---
#title: "Report for Simulations using HiClaSSE model"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
params:
  output_dir: "/output"
---

# HiClaSSE: validation


Here, we validate HiClaSSE against available models from diversitree package.

```{r, setup,  include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here(), cache = TRUE)
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# source dependencies and install them if they are not
source('R/utils/dependencies.R')
#source('R-models/my-ClaSSE.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```
### Make data

```{r}


set.seed(123)
# BiSSE
pars <- c(1.5, 0.5, .1, .1, 1.2, 1)
names(pars) <- diversitree:::default.argnames.bisse()
#pars
phy <- tree.bisse(pars, max.taxa=100, x0=0)
tree <- phy
states <- phy$tip.state
#states
plot(tree)

```



## BiSSE-ness and HiClaSSE

### BiSSE-ness Likelihood


```{r}
##  BiSSEness:
lambda0 = 1.2
p0c = 0.7
p0a = 0.1
lambda1 = 2
p1c = 0.6
p1a = 0.4

# lambda0, lambda1,  mu0, mu1,  q01, q10,   p0c, p0a, p1c, p1a
pars.bi <-c(lambda0, lambda1,  0.03, 0.04,  .5, .6, p0c, p0a,   p1c, p1a)
#pars.bi
lik.bisseness <- make.bisseness(tree, states)
bisseness <- lik.bisseness(pars.bi, intermediates=F, root=ROOT.GIVEN, root.p=c(.5, .5))
#bisseness
```

### HiClaSSE2 Likelihood

Make model

```{r}
Args <- list(
  Nstates = 2L,
  y = list(
    c(0,0, 1,0),
    c(0,0, 0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(2)
#args$pars

```

Parameter mapping

```{r}
lam000 = lambda0 * (1-p0c)
lam001 = lambda0 * p0c*p0a
lam011 = lambda0 * p0c*(1-p0a)
lam100 = lambda1 * p1c*(1-p1a)
lam101 = lambda1 * p1c*p1a
lam111 = lambda1 * (1-p1c)

pars.hi <- c(lam000,lam001,lam011,  lam100,lam101,lam111, 0.03, 0.04,  .5, .6)
names(pars.hi) <- args$pars
print(pars.hi)
```

Inference

```{r}

lik.hiclasse <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
hiclasse <- lik.hiclasse(pars.hi, intermediates=F, root=ROOT.GIVEN, root.p=c(.5, .5), condition.surv=T)

```


### Compare models

The likelihood of BiSSE-ness and HiClaSSE2 are the same

```{r}
print(bisseness)
print(hiclasse)
print(bisseness-hiclasse)
```

## BiSSE and HiClaSSE

### BiSSE

```{r}
##  BiSSEness:
lambda0 = 1.2
lambda1 = 2


# lambda0, lambda1,  mu0, mu1,  q01, q10
pars.bi <-c(lambda0, lambda1,  0.03, 0.04,  .5, .6)
#pars.bi
lik.bisse <- make.bisse(tree, states)
bisse <- lik.bisse(pars.bi, intermediates=F, root=ROOT.GIVEN, root.p=c(.5, .5))
bisse
```


### HiClaSSE2

Make model

```{r}
Args <- list(
  Nstates = 2L,
  y = list(
    c(0,0, 1,0),
    c(0,0, 0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(2)



lam000 = lambda0
lam001 = 0
lam011 = 0
lam100 = 0
lam101 = 0
lam111 = lambda1

pars.hi <- c(lam000,lam001,lam011,  lam100,lam101,lam111, 0.03, 0.04,  0.5, 0.6)
names(pars.hi) <- args$pars
print(pars.hi)

# Inference


lik.hiclasse <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
hiclasse <- lik.hiclasse(pars.hi, intermediates=F, root=ROOT.GIVEN, root.p=c(.5, .5), condition.surv=T)

```


### Compare models

The likelihood is the same

```{r}
print(bisse)
print(hiclasse)
print(bisse-hiclasse)
```




## HiSSE4 and HiClaSSE4


### HiClaSSE4

```{r}

s1 <- 0.6
s2 <- 0.3

sam.fr4 = list(
  c(1-s1, 1-s1, 1-s2, 1-s2,   s1, s1, 0, 0),
  c(1-s1, 1-s1, 1-s2, 1-s2,   0, 0, s2, s2)
)

Args <- list(
  Nstates = 4L,
  y = sam.fr4)

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(4)
#args
#args$pars
length(args$pars)

pars.hc <- c(1:56)
pars.hc[41:56] <- c(1:16)
names(pars.hc) <- args$pars
pars.hc[0:40] <- 0
pars.hc['lam000'] <- 1
pars.hc['lam111'] <- 2
pars.hc['lam222'] <- 3
pars.hc['lam333'] <- 4
pars.hc <- pars.hc/1000
pars.hc
#pars_to_arrays(pars.hc, 4)

#states<- mapvalues(phy$tip.state, from = c("0", "1", "2", "3"), to=c(0, 0, 1, 1) )
lik.hiclasse4 <- make.HiClasse_cpp(phy, states,  sampling.f=NULL,  strict=F, control=list(backend = "gslode"), newArgs)
hiclasse4 <- lik.hiclasse4(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=rep(1/4, 4), condition.surv=T)
#hiclasse4
```



### HiSSE4

```{r}
pars.reoder <- pars_to_arrays(pars.hc, 4)
v=c(1,3,  2,4)
pars.reoder <-reoder_pars(pars.reoder, v)
#pars.reoder
Q_hisse <- TransMatMakerHiSSE(hidden.traits=1)
qs <- extract_off_diagonal(t(pars.reoder$Q))

mu <- pars.reoder$mu
lam <- c(0.001, 0.003, 0.002, 0.004)
div.pars <- convert2ratesHisse(lam, mu)
pars.hisse4 <- c(div.pars$turnover, div.pars$eps, qs)
  
#states<- mapvalues(phy$tip.state, from = c("0", "1", "2", "3"), to=c(0, 0, 1, 1) )
dat <- data.frame(names(phy$tip.state), states)
root <- rep(1/4, 4)

hisse.func <- makeHiSSELikelihood(phy = phy, data = dat, hidden.states = 1, ode.eps=0, root.p=root, f=c(s1,s2))
hisse.func_lik <- hisse.func$log.lik
pars.hisse4 <- setNames(pars.hisse4, names(hisse.func$pars))

lik.hisse <- hisse.func_lik(pars.hisse4)
#lik.hisse



```

### Compare models

The likelihood is the same

```{r}
print(lik.hisse)
print(hiclasse4)
print(lik.hisse-hiclasse4)
```


## Sampling fraction: BiSSE-ness and HiClaSSE

### BiSSE-ness Likelihood


```{r}
##  BiSSEness:
lambda0 = 1.2
p0c = 0.7
p0a = 0.1
lambda1 = 2
p1c = 0.6
p1a = 0.4

# lambda0, lambda1,  mu0, mu1,  q01, q10,   p0c, p0a, p1c, p1a
pars.bi <-c(lambda0, lambda1,  0.03, 0.04,  .5, .6, p0c, p0a,   p1c, p1a)
#pars.bi
lik.bisseness <- make.bisseness(tree, states, sampling.f=c(0.3, 0.6))

#lik.bisseness <- make.bisseness(tree, states, sampling.f=NULL)
bisseness <- lik.bisseness(pars.bi, intermediates=F, root=ROOT.GIVEN, root.p=c(.5, .5))
#bisseness
```

### HiClaSSE2 Likelihood

Make model

```{r}
Args <- list(
  Nstates = 2L,
  y = list(
    c(1-0.3,1-0.6,   0.3,0),
    c(1-0.3,1-0.6,   0,0.6)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(2)
#args$pars

```

Parameter mapping

```{r}
lam000 = lambda0 * (1-p0c)
lam001 = lambda0 * p0c*p0a
lam011 = lambda0 * p0c*(1-p0a)
lam100 = lambda1 * p1c*(1-p1a)
lam101 = lambda1 * p1c*p1a
lam111 = lambda1 * (1-p1c)

pars.hi <- c(lam000,lam001,lam011,  lam100,lam101,lam111, 0.03, 0.04,  .5, .6)
names(pars.hi) <- args$pars
#print(pars.hi)
```

Inference

```{r}

lik.hiclasse <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
hiclasse <- lik.hiclasse(pars.hi, intermediates=F, root=ROOT.GIVEN, root.p=c(.5, .5), condition.surv=T)

```


### Compare models

The likelihood of BiSSE-ness and HiClaSSE2 are the same

```{r}
print(bisseness)
print(hiclasse)
print(bisseness-hiclasse)
```


