---
#title: "BiSSE"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
params:
  output_dir: "/output"
---

# BiSSE congruence (M1-M3)

This example, drawn from Eq. 2 in the main text, demonstrates that the BiSSE2 model with two states (M1) is congruent with BiSSE3 (M2), but is not congruent with BiSSE3 (M3) that violates lumpability.

```{r, include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here(), cache = TRUE)
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# source dependencies and install them if they are not
source('R/utils/dependencies.R')
#source('R-models/my-ClaSSE.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```

Read in data:

```{r}
NSIM = 10
phy <- readRDS(file='R/data/sim/phy_CID4-10tr.RDS')

```


## SET-UP Q matrix for inference

```{r}
Q3 <- make_Qcol(4, c(0.01, 0.01, 0.01, 0))
Q3[1,3] <- Q3[2,1] <- Q3[3,2] <- 0
Q3[4,] <- Q3[4,] <- rep(0, 4)
v=c(1,3,2)
Q3[v,v]
diag(Q3) <- NA
Q3
```



## BiSSE-2 

```{r}

# this is character coding at tips. It sets BiSSE model
Args <- list(
  Nstates = 2L,
  y = list(
    c(0,0, 1,0),
    c(0,0, 0,1)
  ))

newArgs <- makeArgs(Args)
printArgsGlobal()
args <- argnames_HiClaSSE(2)
#args
args$pars
#length(args$pars)

#Q3

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.05
pars.hc['lam111'] <- 0.02
pars.hc['mu0'] <-pars.hc['mu1'] <-0.1
pars.hc['q01'] <- Q3[1,2]
pars.hc['q10'] <- Q3[3,1]

pars.hc
pars_to_arrays(pars.hc, 2)

```

```{r}
args$arrays
```



Run inference

```{r message=FALSE}

BiSSE2 <- list()
# i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/2, 2)
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, lam001~0, lam011~0, lam100~0, lam101~0, mu0 ~ mu1, q01 ~ q10)
  arg.const <- argnames(lik.const)
  starting.point <- pars.hc[arg.const]
  BiSSE2[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}

```


Some helper function to assess the inference parameters
```{r}
# get_item(BiSSE2, 'lnLik')
# pars_to_arrays(BiSSE2[[1]]$par.full, 2)
```


## BiSSE-3     Congruent  
```{r}

Args <- list(
  Nstates = 3L,
  y = list(
    c(0,0,0, 1,0,0),
    c(0,0,0, 0,1,1)
  ))

newArgs <- makeArgs(Args)
printArgsGlobal()
args <- argnames_HiClaSSE(3)
#args
#args$pars
#length(args$pars)

Q30 <- Q3
Q30 <- Q30[-4,-4]
Q30[2,3] <- 0.01
Q30[2,1] <- 0.01
diag(Q30) <- 0
Q30

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.05
pars.hc['lam111'] <- 0.02
pars.hc['lam222'] <- 0.02
pars.hc['mu0'] <-pars.hc['mu1'] <-pars.hc['mu2'] <-0.1
qs <- extract_off_diagonal(Q30)
pars.hc[22:27] <- qs

pars.hc
pars_to_arrays(pars.hc, 3)

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- c(q12 ~ q20, q10 ~ q20, q01 ~ q20)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0)
f.lams <- c(lam222 ~  lam111)
f.list<- c(zero.constr,  f.lams, f.mu, f.qs)


```

 Run inference
 
```{r message=FALSE}

BiSSE3C <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/2, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  starting.point <- pars.hc[arg.const]
  BiSSE3C[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```


Some helper function to assess the inference parameters
```{r}
# get_item(BiSSE3C, 'lnLik')
# #get_item(BiSSE2, 'lnLik')
# pars_to_arrays(BiSSE3C[[1]]$par.full, 3)
```


## BiSSE-3 Congruent, Inferred Pars  

Mapping of parameters between congruent models is deterministic. Here, we derive the parameters directly from BiSSE2 model.

```{r message=FALSE}
BiSSE3C_pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #---
  mi2_est <- get_item(BiSSE2[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- mi2_est['lam000']
  pars.hc['lam111'] <- mi2_est['lam111']
  pars.hc['lam222'] <- mi2_est['lam111']
  pars.hc['mu0'] <-pars.hc['mu1'] <-pars.hc['mu2'] <- mi2_est['mu1']
  qs <- extract_off_diagonal(Q30)
  pars.hc[22:27] <- c(mi2_est['q10'], 0, mi2_est['q10'], mi2_est['q10'], mi2_est['q10'], 0)
  #pars.hc
  #pars_to_arrays(pars.hc, 3)
  #---
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/2, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  BiSSE3C_pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```



## BiSSE-3     Non-congruent    

```{r}

Args <- list(
  Nstates = 3L,
  y = list(
    c(0,0,0, 1,0,0),
    c(0,0,0, 0,1,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(3)
# args
# args$pars
# length(args$pars)

Q30 <- Q3
Q30 <- Q30[-4,-4]
Q30[2,3] <- 0.01
diag(Q30) <- 0
Q30

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.05
pars.hc['lam111'] <- 0.02
pars.hc['lam222'] <- 0.02
pars.hc['mu0'] <-pars.hc['mu1'] <-pars.hc['mu2'] <-0.1
qs <- extract_off_diagonal(Q30)
pars.hc[22:27] <- qs

# pars.hc
# pars_to_arrays(pars.hc, 3)

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- c(q12 ~ q20, q01 ~ q20)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0)
f.lams <- c(lam222 ~  lam111)
f.list<- c(zero.constr,  f.lams, f.mu, f.qs)

```

Inference

```{r message=FALSE}

BiSSE3 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/2, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  #res <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  starting.point <- pars.hc[arg.const]
  BiSSE3[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}

```



## Another Non-congruent Models


If lumpability does not hold, then shuffling parameters within partition subsets results in different models. Below, M4 is identical to BiSSE4-NonCongruent, and M5 represents another distinct model where 'lam112' is set to 0.2.

```{r message=FALSE}
Qnc <- c(0.05, 0, 0.05, 0.05, 0, 0)
M4 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #---
  #mi2_est <- get_item(BiSSE2[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- 0.1
  pars.hc['lam111'] <- 0.2
  pars.hc['lam222'] <- 0.2
  pars.hc['mu0'] <-pars.hc['mu1'] <-pars.hc['mu2'] <- 0.001
  #qs <- extract_off_diagonal(Q30)
  pars.hc[22:27] <- Qnc
  #pars.hc
  #pars_to_arrays(pars.hc, 3)
  #---
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/2, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  M4[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

M5 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #---
  #mi2_est <- get_item(BiSSE2[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- 0.1
  pars.hc['lam112'] <- 0.2
  pars.hc['lam222'] <- 0.2
  pars.hc['mu0'] <-pars.hc['mu1'] <-pars.hc['mu2'] <- 0.001
  #qs <- extract_off_diagonal(Q30)
  pars.hc[22:27] <- Qnc
  #pars.hc
  #pars_to_arrays(pars.hc, 3)
  #---
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/2, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  M5[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}


unlist(M4)
unlist(M5)

```

## Zero Extinction

```{r message=FALSE}

Args <- list(
  Nstates = 3L,
  y = list(
    c(0,0,0, 1,0,0),
    c(0,0,0, 0,1,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(3)
#args
#args$pars
length(args$pars)

Qnc <- c(0.05, 0, 0.05, 0.05, 0.05, 0)
M4 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #---
  #mi2_est <- get_item(BiSSE2[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- 0.1
  pars.hc['lam111'] <- 0.2
  pars.hc['lam222'] <- 0.2
  pars.hc['mu0'] <-pars.hc['mu1'] <-pars.hc['mu2'] <- 0
  #qs <- extract_off_diagonal(Q30)
  pars.hc[22:27] <- Qnc
  #pars.hc
  #pars_to_arrays(pars.hc, 3)
  #---
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/2, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  M4[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

M5 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #---
  #mi2_est <- get_item(BiSSE2[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- 0.1
  pars.hc['lam100'] <- 0.2
  pars.hc['lam222'] <- 0.2
  pars.hc['mu0'] <-pars.hc['mu1'] <-pars.hc['mu2'] <- 0
  #qs <- extract_off_diagonal(Q30)
  pars.hc[22:27] <- Qnc
  #pars.hc
  #pars_to_arrays(pars.hc, 3)
  #---
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/2, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  M5[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}


unlist(M4)
unlist(M5)

```

## Results 

### Likelihood table

The likelihood scores of the used models.

```{r}


mi2 = get_item(BiSSE2, 'lnLik')
mi3c = get_item(BiSSE3C, 'lnLik')
mi3c_pars = unlist(BiSSE3C_pars)
mi3 = get_item(BiSSE3, 'lnLik')

BiSSE <- tibble(
"N" = c(1:10),
"BiSSE2" = mi2,
"congruent BiSSE3 (BiSSE2)"=mi3c,
"congruent BiSSE3 (pars of BiSSE2)"= mi3c_pars,
"Non-congruent BiSSE3" = mi3)

kable(BiSSE, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Mean absolute error


The table and plot below demonstrate that the likelihoods of  BiSSE2 and BiSSE3-Congruent are the same; the differences arise solely from numerical integration. In constrast, the differences between BiSSE2 and BiSSE3-NonCongruent are substantial, indicating they are distinct models. 

```{r}
d1 =abs(mi2-mi3c) %>% mean()
d2 =abs(mi2-mi3c_pars) %>% mean()
d3 = abs(mi2-mi3) %>% mean()

# Data
delta_values <- tibble(
  Delta = c("mi2-mi3_Cong", "mi2-mi3_Cong_pars", "mi2-mi3_nonCongruent"),
  Value = c(d1, d2, d3)
)

kable(delta_values, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Create the plot
ggplot(delta_values, aes(x = Delta, y = Value, label = sprintf("%.2e", Value))) +
  geom_col(fill = c("blue", "green", "red")) +
  geom_text(vjust = -0.5, angle = 45) +
  labs(title = "Comparison of Ln Differences", x = "Deltas", y = "Magnitude") +
  theme_minimal()



```

### Parameter estimates 

```{r}

get_item(BiSSE2[1], 'par')
get_item(BiSSE3C[1], 'par')
get_item(BiSSE3[1], 'par')

# pars_to_arrays(BiSSE2[[1]]$par.full, 2)
# pars_to_arrays(BiSSE3C[[1]]$par.full, 3)
# pars_to_arrays(BiSSE3[[1]]$par.full, 3)
```

