---
#title: " Congruence between independent and dependent SSEs"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
params:
  output_dir: "/output"
---

# Congruence between independent and dependent SSEs

| Model    | np | pars                           | Cong. with CID4 | Trait & Div.     | Speciation Rates              |
|----------|----|--------------------------------|---------------------|------------------|--------------------------------|
| CID4     | 4  | 2d 1q 1mu                      | --                  | independent      | decoupled (d)                 |
| CID8     | 4  | 2d 1q 1mu                      | yes                 | independent      | decoupled (d)                 |
| COR8-C   | 4  | 2d 1q 1mu                      | yes                 | correlated       | decoupled (d)                 |
| CLA8-C   | 4  | 2c 1q 1mu                      | yes                 | correlated       | coupled (c)                   |
| COR8-NC  | 4  | 2d 1q 1mu                      | no                  | correlated*      | decoupled (d)                 |
| CLA8-NC  | 4  | 2c 1q 1mu                      | no                  | correlated       | coupled* (c)                  |

*indicates non-lumpable transition or speciation rates. See also, Fig. referenced and Table S1.


```{r,  include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here(), cache = TRUE)
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
source('R/utils/dependencies.R')
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```

## Read in data

```{r}
NSIM = 10
phy <- readRDS(file='R/data/phy_CID4.RDS')
```

## Setting-up Q matrices for inference

### CID4

```{r}
# trait
Q_t <- initQ(c(0,1), c(.1, .1))

# diversification regime
Q_r2 <- initQ(c('A','B'), c(.1,.1))


# speciation rates for tracking the order
La4 <- diag(c(-0.1,-0.05),2)
La4 = La4 %x% diag(1,2)
# order according to regimes
Q_cid4.r <- amaSMM(Q_r2, Q_t)

# order according to trait
v=c(1,3, 2,4)
Q_cid4.t <- Q_cid4.r[v,v]

print(Q_cid4.r)
print(La4)
print(Q_cid4.t)
print(La4[v,v])

```


### CID8 (congruent to CID4)

```{r}
# diversification regime
#Q_r4 <- initQ(c('a','b', 'c', 'd'), c(.01,.01))

Q_r4 <-amaSMM(Q_t, Q_t) 
colnames(Q_r4) <- rownames(Q_r4) <- c('a','b', 'c', 'd')

# speciation rates for tracking the order
La8 <- diag(c(-0.1,-0.05),2)
La8 = La8 %x% diag(1,4)

# order according to regimes
Q_cid8.r <- amaSMM(Q_r4, Q_t) 

# order according to trait
v=c(1,3,5,7, 2,4,6,8)
Q_cid8.t <- Q_cid8.r[v,v]

print(Q_cid8.r)
print(La8)
print(Q_cid8.t)
print(La8[v,v])

```

### COR8-C (correlated evolution, congruent to CID4)

```{r}
# like CID8 but with correlation
Q_cor8_C.t <- Q_cid8.t

bl <- matrix(c(0,.1,.1,0),2,2)
Q_cor8_C.t[1:2,5:6] <- bl
Q_cor8_C.t[3:4,7:8] <- bl
Q_cor8_C.t[5:6,1:2] <- bl
Q_cor8_C.t[7:8,3:4] <- bl

v=c(1,5, 2,6, 3,7, 4,8)
Q_cor8_C.r <- Q_cor8_C.t[v,v]

print(Q_cor8_C.r)
#print(La8)
print(Q_cor8_C.t)
#print(La8[v,v])

```

### COR8-NC (Non-correlated evolution, congruent to CID4)

```{r}
# like CID8 but not congruent
Q_cor8_NC.t <- Q_cid8.t

bl <- matrix(c(0,0,.1,0),2,2)
Q_cor8_NC.t[1:2,5:6] <- bl
Q_cor8_NC.t[3:4,7:8] <- bl
Q_cor8_NC.t[5:6,1:2] <- bl
Q_cor8_NC.t[7:8,3:4] <- bl

v=c(1,5, 2,6, 3,7, 4,8)
Q_cor8_NC.r <- Q_cor8_NC.t[v,v]

print(Q_cor8_NC.r)
#print(La8)
print(Q_cor8_NC.t)
#print(La8[v,v])

```

## Irreducible model

### CID4 (4 pars)

```{r}


Args <- list(
  Nstates = 4L,
  y = list(
    c(0,0,0,0, 1,0,1,0),
    c(0,0,0,0, 0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(4)
#args$arrays
#args$pars
#length(args$pars)
reoder_lambdas(args$arrays, c(1,3, 2,4))

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.05
pars.hc['lam333'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <-0.1
qs <- extract_off_diagonal(Q_cid4.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,4)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid4.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam333 ~ lam222)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r message=FALSE}

CID4 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/4, 1/4, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  starting.point <- pars.hc[arg.const]
  CID4[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```

Some helper functions to assess the inference parameters.

```{r}
get_item(CID4, 'lnLik')
pars_to_arrays(CID4[[1]]$par.full, 4) %>% reoder_lambdas(., c(1,3, 2,4))
get_item(CID4[1], 'par')
```

## Inferences using parameters from CID4

### CID8 (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
args$arrays
#reoder_lambdas(args$arrays, c(1,3,5,7, 2,4,6,8))
```

Run inference

```{r message=FALSE}


CID8.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- est['lam000']
  pars.hc['lam111'] <- est['lam000']
  pars.hc['lam222'] <- est['lam000']
  pars.hc['lam333'] <- est['lam000']
  pars.hc['lam444'] <- est['lam222']
  pars.hc['lam555'] <- est['lam222']
  pars.hc['lam666'] <- est['lam222']
  pars.hc['lam777'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cid8.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  #pars.hc
  #pars_to_arrays(pars.hc, 8)
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  CID8.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Compare Ln

```{r}
get_item(CID4, 'lnLik')
CID8.pars %>% unlist

pars_to_arrays(pars.hc, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```

### COR8-C (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args

```

Run inference

```{r message=FALSE}


COR8_C.pars <- list()
i=2
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- est['lam000']
  pars.hc['lam111'] <- est['lam000']
  pars.hc['lam222'] <- est['lam000']
  pars.hc['lam333'] <- est['lam000']
  pars.hc['lam444'] <- est['lam222']
  pars.hc['lam555'] <- est['lam222']
  pars.hc['lam666'] <- est['lam222']
  pars.hc['lam777'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_C.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  #v=c(1,3,5,7, 2,4,6,8)
  #pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  COR8_C.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Compare Ln

```{r}
get_item(CID4, 'lnLik')
COR8_C.pars %>% unlist

pars_to_arrays(pars.hc, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```

### COR8-NC (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args

```

Run inference

```{r message=FALSE}


COR8_NC.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- est['lam000']
  pars.hc['lam111'] <- est['lam000']
  pars.hc['lam222'] <- est['lam000']
  pars.hc['lam333'] <- est['lam000']
  pars.hc['lam444'] <- est['lam222']
  pars.hc['lam555'] <- est['lam222']
  pars.hc['lam666'] <- est['lam222']
  pars.hc['lam777'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_NC.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  # v=c(1,3,5,7, 2,4,6,8)
  # pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  COR8_NC.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Some helper function to assess the inference parameters

```{r}
get_item(CID4, 'lnLik')
COR8_NC.pars %>% unlist

pars_to_arrays(pars.hc, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```

### CLA8-C (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
# reoder_lambdas(args$arrays, c(1,3,5,7, 2,4,6,8))
```

Run inference

```{r message=FALSE}


CLA8_C.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  # pars.hc['lam000'] <- est['lam000']
  # pars.hc['lam111'] <- est['lam000']
  # pars.hc['lam222'] <- est['lam000']
  # pars.hc['lam333'] <- est['lam000']
  # pars.hc['lam444'] <- est['lam222']
  # pars.hc['lam555'] <- est['lam222']
  # pars.hc['lam666'] <- est['lam222']
  # pars.hc['lam777'] <- est['lam222']
  pars.hc['lam022'] <- est['lam000']
  pars.hc['lam133'] <- est['lam000']
  pars.hc['lam202'] <- est['lam000']
  pars.hc['lam311'] <- est['lam000']
  pars.hc['lam466'] <- est['lam222']
  pars.hc['lam577'] <- est['lam222']
  pars.hc['lam646'] <- est['lam222']
  pars.hc['lam757'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_C.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  # v=c(1,3,5,7, 2,4,6,8)
  # pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  CLA8_C.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Some helper function to assess the inference parameters

```{r}
get_item(CID4, 'lnLik')
CLA8_C.pars %>% unlist
pars_to_arrays(pars.hc, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```

### CLA8-NC (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#v=c(1,3,5,7, 2,4,6,8)
#reoder_lambdas(args$arrays, c(1,3,5,7, 2,4,6,8))
```

Run inference

```{r message=FALSE}


CLA8_NC.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  # lam004, lam116, lam422 violate lumpability
  pars.hc['lam004'] <- est['lam000']
  pars.hc['lam116'] <- est['lam000']
  pars.hc['lam202'] <- est['lam000']
  pars.hc['lam311'] <- est['lam000']
  pars.hc['lam422'] <- est['lam222']
  pars.hc['lam577'] <- est['lam222']
  pars.hc['lam646'] <- est['lam222']
  pars.hc['lam757'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_C.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  # v=c(1,3,5,7, 2,4,6,8)
  # pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  CLA8_NC.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Some helper function to assess the inference parameters

```{r}
get_item(CID4, 'lnLik')
CLA8_NC.pars %>% unlist
pars_to_arrays(pars.hc, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```

## Maximum Likelihood. All models have 4 pars.

### CID8

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.1
pars.hc['lam333'] <- 0.1
pars.hc['lam444'] <- 0.05
pars.hc['lam555'] <- 0.05
pars.hc['lam666'] <- 0.05
pars.hc['lam777'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cid8.r)
qsl=length(qs)
#unique(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid8.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam222 ~ lam000, lam333 ~ lam000,   
            lam555 ~ lam444, lam666 ~ lam444, lam777 ~ lam444)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r}

CID8 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  starting.point <- pars.hc[arg.const]
  # starting.point <- get_item(CID4[i], 'par')
  # names(starting.point) <- c('lam444', 'lam000', 'mu0', 'q01', 'q02')
  # starting.point['q01'] <- starting.point['q02']/2
  
  CID8[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```

Assessing the inferred parameters

```{r}
get_item(CID4, 'lnLik')
get_item(CID8, 'lnLik')

get_item(CID8[1], 'par')
get_item(CID4[1], 'par')

#pars_to_arrays(CID8[[1]]$par.full, 8)
pars_to_arrays(CID8[[1]]$par.full, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```




### COR8-C

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.1
pars.hc['lam333'] <- 0.1
pars.hc['lam444'] <- 0.05
pars.hc['lam555'] <- 0.05
pars.hc['lam666'] <- 0.05
pars.hc['lam777'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cor8_C.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cor8_C.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam222 ~ lam000, lam333 ~ lam000,   
            lam555 ~ lam444, lam666 ~ lam444, lam777 ~ lam444)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r}

COR8_C <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  starting.point <- pars.hc[arg.const]
  # starting.point <- get_item(CID8[i], 'par')
  # names(starting.point) <- arg.const
  # starting.point['q02'] <- get_item(CID8[i], 'par')['q02']
  # starting.point['q03'] <- get_item(CID8[i], 'par')['q01']
  COR8_C[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```

Assessing the inferred parameters

```{r}
get_item(CID4, 'lnLik')
get_item(CID8, 'lnLik')
get_item(COR8_C, 'lnLik')

get_item(COR8_C[1], 'par')

pars_to_arrays(COR8_C[[1]]$par.full, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))

```


### COR8-NC

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.1
pars.hc['lam333'] <- 0.1
pars.hc['lam444'] <- 0.05
pars.hc['lam555'] <- 0.05
pars.hc['lam666'] <- 0.05
pars.hc['lam777'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cor8_NC.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cor8_NC.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam222 ~ lam000, lam333 ~ lam000,   
            lam555 ~ lam444, lam666 ~ lam444, lam777 ~ lam444)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r}

COR8_NC <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  # starting.point <- pars.hc[arg.const]
  starting.point <- get_item(COR8_C[i], 'par')
  COR8_NC[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```

Assessing the inferred parameters

```{r}
get_item(CID4, 'lnLik')
get_item(CID8, 'lnLik')
get_item(COR8_NC, 'lnLik')

get_item(CID8[1], 'par')

pars_to_arrays(COR8_NC[[1]]$par.full, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```




### CLA8-C

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
# pars.hc['lam000'] <- 0.1
# pars.hc['lam111'] <- 0.1
# pars.hc['lam222'] <- 0.1
# pars.hc['lam333'] <- 0.1
# pars.hc['lam444'] <- 0.05
# pars.hc['lam555'] <- 0.05
# pars.hc['lam666'] <- 0.05
# pars.hc['lam777'] <- 0.05
pars.hc['lam022'] <- 0.1
pars.hc['lam133'] <- 0.1
pars.hc['lam202'] <- 0.1
pars.hc['lam311'] <- 0.1
pars.hc['lam466'] <- 0.05
pars.hc['lam577'] <- 0.05
pars.hc['lam646'] <- 0.05
pars.hc['lam757'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cor8_C.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cor8_C.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam133 ~ lam022, lam202 ~ lam022, lam311 ~ lam022,   
            lam577 ~ lam466, lam646 ~ lam466, lam757 ~ lam466)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r}

CLA8_C <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  #starting.point <- pars.hc[arg.const]
  starting.point <- get_item(COR8_C[i], 'par')
  names(starting.point) <- arg.const
  CLA8_C[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```

Assessing the inferred parameters

```{r}
get_item(CID4, 'lnLik')
get_item(CLA8_C, 'lnLik')

get_item(CID8[1], 'par')

pars_to_arrays(CLA8_C[[1]]$par.full, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```




### CLA8-NC

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
# pars.hc['lam000'] <- 0.1
# pars.hc['lam111'] <- 0.1
# pars.hc['lam222'] <- 0.1
# pars.hc['lam333'] <- 0.1
# pars.hc['lam444'] <- 0.05
# pars.hc['lam555'] <- 0.05
# pars.hc['lam666'] <- 0.05
# pars.hc['lam777'] <- 0.05
# lam004, lam116, lam422 violate lumpability
pars.hc['lam004'] <- 0.1
pars.hc['lam116'] <- 0.1
pars.hc['lam202'] <- 0.1
pars.hc['lam311'] <- 0.1
pars.hc['lam422'] <- 0.05
pars.hc['lam577'] <- 0.05
pars.hc['lam646'] <- 0.05
pars.hc['lam757'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cor8_C.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cor8_C.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam116 ~ lam004, lam202 ~ lam004, lam311 ~ lam004,   
            lam577 ~ lam422, lam646 ~ lam422, lam757 ~ lam422)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r}

CLA8_NC <- list()
i=1
# skip 4 as mle search gets stuck and R crushes
seq <- 1:10
seq <-seq[-4]
for (i in seq){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  #starting.point <- pars.hc[arg.const]
  starting.point <- get_item(CLA8_C[i], 'par')
  names(starting.point) <- arg.const
  CLA8_NC[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```

Assessing the inferred parameters

```{r}
get_item(CID4, 'lnLik')
get_item(COR8_NC, 'lnLik')
get_item(CLA8_NC, 'lnLik')

get_item(CID8[1], 'par')

pars_to_arrays(CLA8_C[[1]]$par.full, 8) %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```


## Results 

### Inferences using parameters from CID4

#### Likelihood table

```{r}


cid4 = get_item(CID4, 'lnLik')

cor.pars <- tibble(
"N" = c(1:NSIM),
"cid4" = cid4,
"CID8.pars"= CID8.pars %>% unlist,
"COR8-C.pars"= COR8_C.pars %>% unlist,
"CLA8-C.pars" = CLA8_C.pars %>% unlist,
"COR8-NC.pars" = COR8_NC.pars %>% unlist,
"CLA8-NC.pars" = CLA8_NC.pars %>% unlist,
)

kable(cor.pars, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


#### Mean absolute error

```{r}

dcor.pars <- tibble(
"cid4" = cid4,
"CID8.pars"= CID8.pars %>% unlist,
"COR8-C.pars"= COR8_C.pars %>% unlist,
"CLA8-C.pars" = CLA8_C.pars %>% unlist,
"COR8-NC.pars" = COR8_NC.pars %>% unlist,
"CLA8-NC.pars" = CLA8_NC.pars %>% unlist,
)
cid4t <- cid4
dcor.pars <- dcor.pars %>%
  mutate_all(~ abs(cid4t - .) )

column_means <- dcor.pars %>%
  summarise_all(~ mean(.))

# Data
delta_values <- tibble(
  Delta = names(column_means),
  Value = column_means %>% slice(1) %>% unlist()
)

kable(column_means, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Create the plot
ggplot(delta_values, aes(x = Delta, y = Value, label = sprintf("%.2e", Value))) +
  geom_col(fill = c("blue", "blue", "blue", "blue", "red",  "red")) +
  geom_text(vjust = -0.5, angle = 45) +
  labs(title = "Comparison of Ln Differences", x = "Deltas", y = "Magnitude") +
  theme_minimal()


```

### Maximum Likelihood estimates

#### Likelihood table

```{r}

cor.pars <- tibble(
"N" = c(1:NSIM),
"cid4" = cid4,
"CID8"= get_item(CID8, 'lnLik'),
"COR8-C"= get_item(COR8_C, 'lnLik'),
"CLA8-C" = get_item(CLA8_C, 'lnLik'),
"COR8-NC" = get_item(COR8_NC, 'lnLik'),
"CLA8-NC" = append(get_item(CLA8_NC, 'lnLik'), NA, after=3),
)

kable(cor.pars, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


#### Mean absolute error

```{r}

dcor.pars <- tibble(
"cid4" = cid4,
"CID8"= get_item(CID8, 'lnLik'),
"COR8-C"= get_item(COR8_C, 'lnLik'),
"CLA8-C" = get_item(CLA8_C, 'lnLik'),
"COR8-NC" = get_item(COR8_NC, 'lnLik'),
"CLA8-NC" = append(get_item(CLA8_NC, 'lnLik'), NA, after=3),
)

# dcor.pars <- tibble(
# "cid4" = cid4,
# "CID8.pars"= CID8.pars %>% unlist,
# "COR8-C.pars"= COR8_C.pars %>% unlist,
# "CLA8-C.pars" = CLA8_C.pars %>% unlist,
# "COR8-NC.pars" = COR8_NC.pars %>% unlist,
# "CLA8-NC.pars" = CLA8_NC.pars %>% unlist,
# )
cid4t <- cid4
dcor.pars <- dcor.pars %>%
  mutate_all(~ abs(cid4t - .) )

column_means <- dcor.pars %>%
  summarise_all(~ mean(., na.rm = TRUE))

# Data
delta_values <- tibble(
  Delta = names(column_means),
  Value = column_means %>% slice(1) %>% unlist()
)

kable(column_means, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Create the plot
ggplot(delta_values, aes(x = Delta, y = Value, label = sprintf("%.2e", Value))) +
  geom_col(fill = c("blue", "blue", "blue", "blue", "red",  "red")) +
  geom_text(vjust = -0.5, angle = 45) +
  labs(title = "Comparison of Ln Differences", x = "Deltas", y = "Magnitude") +
  theme_minimal()


```


