---
#title: "EHE"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
params:
  output_dir: "/output"
cache: TRUE
---

# Equal Rate Hidden Expansion (EHE)

HE can be extended to a point when all permissible rates are equal, except extinctions. We refer to this transformation as Equal Rate Hidden Expansion (EHE).

The EHE model can be created for any SSE, if we know parameter values of that SSE. In this example, we calculate likelihood for CID4 with known parameters and construct congruent EHE8-C model. We show that the likelihoods are identical up to numerical integration error.

```{r, include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here(), cache = TRUE)
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
source('R/utils/dependencies.R')
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```

## Read in data

```{r}
NSIM = 10
phy <- readRDS(file='R/data/phy_CID4.RDS')
```

## Setting-up Q matrices for inference

### CID4

```{r}
# trait
Q_t <- initQ(c(0,1), c(2, 2))
# diversification regime
Q_r2 <- initQ(c('A','B'), c(2,2))
# speciation rates for tracking the order
La4 <- diag(c(-3,-1),2)
La4 = La4 %x% diag(1,2)
# order according to regimes
Q_cid4.r <- amaSMM(Q_r2, Q_t)

# order according to trait
v=c(1,3, 2,4)
Q_cid4.t <- Q_cid4.r[v,v]

print(Q_cid4.r)
print(La4)
print(Q_cid4.t)
print(La4[v,v])


```

### CID8 (congruent to CID4)

```{r}
# diversification regime
# Q_r4 <- initQ(c('a','b', 'c', 'd'), c(2,2))
Q_r4 <- amaSMM(Q_r2, Q_r2)
colnames(Q_r4) <- rownames(Q_r4) <- c('a','b', 'c', 'd')

# speciation rates for tracking the order
La8 <- diag(c(-3,-1),2)
La8 = La8 %x% diag(1,4)

# order according to regimes
Q_cid8.r <- amaSMM(Q_r4, Q_t) 

# order according to trait
v=c(1,3,5,7, 2,4,6,8)
Q_cid8.t <- Q_cid8.r[v,v]

print(Q_cid8.r)
print(La8)
print(Q_cid8.t)
print(La8[v,v])

```

### EHE8-C (correlated evolution, congruent to CID4)

```{r}

# Create the matrix as a data frame
Q_ehe8_C.t <- data.frame(
  a0 = c(-4, 0, 1, 1, 1, 1, 0, 0),
  b0 = c(0, -4, 1, 1, 1, 1, 0, 0),
  c0 = c(1, 1, -4, 0, 0, 0, 1, 1),
  d0 = c(1, 1, 0, -4, 0, 0, 1, 1),
  a1 = c(1, 1, 0, 0, -4, 0, 1, 1),
  b1 = c(1, 1, 0, 0, 0, -4, 1, 1),
  c1 = c(0, 0, 1, 1, 1, 1, -4, 0),
  d1 = c(0, 0, 1, 1, 1, 1, 0, -4)
)

# Set row names and column names
rownames(Q_ehe8_C.t) <- colnames(Q_ehe8_C.t)
Q_ehe8_C.t <-as.matrix(Q_ehe8_C.t)

v=c(1,5, 2,6, 3,7, 4,8)
Q_ehe8_C.r <- Q_ehe8_C.t[v,v]

print(Q_ehe8_C.r)
#print(La8)
print(Q_ehe8_C.t)
print(La8[v,v])

```


## Inferences using parameters from CID4

### CID4 (4 pars)


We assume that we know the parameter estimated for CID4
```{r}
lam1=0.3
lam2=0.1
q=0.2
mu=0.01
```




Calculating likelihood

```{r message=FALSE}

Args <- list(
  Nstates = 4L,
  y = list(
    c(0,0,0,0, 1,0,1,0),
    c(0,0,0,0, 0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(4)

CID4.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- lam1
  pars.hc['lam111'] <- lam1
  pars.hc['lam222'] <- lam2
  pars.hc['lam333'] <- lam2
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3']  <- mu
  qs <- extract_off_diagonal(Q_cid4.r/10)
  qsl=length(qs)
  #qs[qs==1] <- est['q01']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  #pars.hc
  #pars_to_arrays(pars.hc, 4)
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/4, 4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  CID4.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

This is CID4 model.

```{r}
#pars_to_arrays(pars.hc, 4) %>% reoder_lambdas(., c(1,3, 2,4))
```


### EHE8-C (2 pars)

Now we create EHE8-C by decomposing our known parameter estimates into just two parameters: (1.) speciation and transitions rates equal to 0.1; and extinction equal to 0.01.


```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#reoder_lambdas(args$arrays, c(1,3,5,7, 2,4,6,8))

```

Let's check the model rates
```{r}
# a0 a1 b0 b1 c0 c1 d0 d1
ar=names2array(args$arrays, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))
ar.re=reoder_lambdas(ar, c(1,3,5,7, 2,4,6,8))
#ar.re$lam.tensor


# rates that will be used to setup model
element <- c("lam000", 'lam002', 'lam022', 
             'lam200', 'lam202', 'lam222', 
             'lam111', 'lam113', 'lam133',
             'lam311', 'lam313', 'lam333',
             'lam444', 'lam646', 'lam555', 'lam777'
             )
find_elements(ar.re$lam.tensor, element)
```




Calculating Ln

```{r message=FALSE}

lam1=0.3
lam2=0.1
q=0.2
mu=0.01

EHE8_C.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  # pars.hc['lam000'] <- lam1
  # pars.hc['lam111'] <- lam1
  # pars.hc['lam222'] <- lam1
  # pars.hc['lam333'] <- lam1
  pars.hc['lam000'] <- pars.hc['lam002'] <- pars.hc['lam022'] <- lam1/3
  pars.hc['lam200'] <- pars.hc['lam202'] <- pars.hc['lam222'] <- lam1/3
  pars.hc['lam111'] <- pars.hc['lam113'] <- pars.hc['lam133'] <- lam1/3
  pars.hc['lam311'] <- pars.hc['lam313'] <- pars.hc['lam333'] <- lam1/3
  pars.hc['lam444'] <- lam2
  pars.hc['lam555'] <- lam2
  #pars.hc['lam666'] <- lam2
  pars.hc['lam646'] <- lam2 # this makes the model trait-dependent
  pars.hc['lam777'] <- lam2
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- mu
  qs <- extract_off_diagonal(Q_ehe8_C.r/10)
  qsl=length(qs)
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  #pars.hc
  pars_to_arrays(pars.hc, 8)
  reoder_lambdas(pars_to_arrays(pars.hc, 8), c(1,3,5,7, 2,4,6,8))
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  EHE8_C.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```


#### EHE: show model topology

```{r}
# Plot the graph
ar <- pars_to_arrays(pars.hc, 8)
Qpl <- names2array(ar, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))$Q
g <- create_graph_from_matrix(Qpl)
unidirectional_edges <- is_unidirectional(g)
plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))

```
```{r}
ar <- pars_to_arrays(pars.hc, 8)
ar=names2array(ar, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))
ar.re=reoder_lambdas(ar, c(1,3,5,7, 2,4,6,8))
ar.re
```


## Compare Ln estimates

The likelihoods are the same, the differences are due to numerical integration.

```{r}
cid4 <- CID4.pars %>% unlist
print(cid4)
ehe8 <- EHE8_C.pars %>% unlist
print(ehe8)
print(cid4-ehe8)

# > print(cid4-ehe8)
#  [1] 2.922985e-06 2.760492e-06 3.440786e-06 3.040528e-06 2.149453e-06 3.493703e-06 2.752414e-06 2.570285e-06 2.996501e-06
# [10] 3.601493e-06
```



