---
#title: "NON-Semi-Congruent"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
params:
  output_dir: "/output"
cache: TRUE
---

# SEM: semicongruent models

In this example, we data simulated under the CID4 model for three different scenarios: 100, 500, and 1000 tips. Subsequently, we conduct ML inference using two models: the original CID4 and the semi-congruent EHE8-C model. The CID4 model has four parameters, while the EHE8-C model has only two.

We observe that the likelihood of the CID4 model is always better than that of the EHE8-C model across all scenarios. However, due to fewer parameters, the AIC of the EHE8-C tends to be better in most trials.

It's worth noting that lumping the semi-congruent EHE8-C model produces the CID4 model. In other words, if we create an irreducible model from the EHE8-C model, we obtain the original CID4 model with its four parameters.

```{r,  include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
source('R/utils/dependencies.R')
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```

## Setting-up Q matrices for inference

### CID4

```{r}
# trait
Q_t <- initQ(c(0,1), c(2, 2))
# diversification regime
Q_r2 <- initQ(c('A','B'), c(2,2))
# speciation rates for tracking the order
La4 <- diag(c(-3,-1),2)
La4 = La4 %x% diag(1,2)
# order according to regimes
Q_cid4.r <- amaSMM(Q_r2, Q_t)

# order according to trait
v=c(1,3, 2,4)
Q_cid4.t <- Q_cid4.r[v,v]

print(Q_cid4.r)
print(La4)
print(Q_cid4.t)
print(La4[v,v])


```

### SEM8

Similar to EHE model.

```{r}

# Create the matrix as a data frame
Q_ehe8_C.t <- data.frame(
  a0 = c(-4, 0, 1, 1, 1, 1, 0, 0),
  b0 = c(0, -4, 1, 1, 1, 1, 0, 0),
  c0 = c(1, 1, -4, 0, 0, 0, 1, 1),
  d0 = c(1, 1, 0, -4, 0, 0, 1, 1),
  a1 = c(1, 1, 0, 0, -4, 0, 1, 1),
  b1 = c(1, 1, 0, 0, 0, -4, 1, 1),
  c1 = c(0, 0, 1, 1, 1, 1, -4, 0),
  d1 = c(0, 0, 1, 1, 1, 1, 0, -4)
)

# Set row names and column names
rownames(Q_ehe8_C.t) <- colnames(Q_ehe8_C.t)
Q_ehe8_C.t <-as.matrix(Q_ehe8_C.t)


v=c(1,5, 2,6, 3,7, 4,8)
Q_ehe8_C.r <- Q_ehe8_C.t[v,v]

print(Q_ehe8_C.r)
#print(La8)
print(Q_ehe8_C.t)
#print(La8[v,v])

```


## Read in data

```{r}

NSIM=100
files_base <- c('phy_CID4-100tips-100tr', 'phy_CID4-500tips-100tr', 'phy_CID4-1000tips-100tr')

```

## ML: CID4 (4 pars)

```{r}


Args <- list(
  Nstates = 4L,
  y = list(
    c(0,0,0,0, 1,0,1,0),
    c(0,0,0,0, 0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(4)
#args$arrays
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.05
pars.hc['lam333'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <-0.1
qs <- extract_off_diagonal(Q_cid4.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,4)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid4.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam333 ~ lam222)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r eval=FALSE}
#base="phy_CID4-100tips-100tr"
for (base in files_base[1:3]){
  file=file.path("R/data/sim", paste0(base, '.RDS'))
  print(paste0('Reading file: ',file))
  phy <- readRDS(file=file)

  CID4 <- list()
  #i=1
  #which(is.na(phy))
  for (i in 1:NSIM){

    print(paste0('Working on: ', i))
    tree <- phy[[i]]
    states<- tree$tip.state
    states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
    root <- c(1/4, 1/4, 1/4, 1/4)

    lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
    lik.const <- constrain(lik.c, formulae = f.list)
    arg.const <- argnames(lik.const)
    starting.point <- pars.hc[arg.const]
    CID4[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
    #CID4[[1]]$lnLik

  }
  file_out=paste0('R/data/out/lik-CID4-', base, '.RDS')
  saveRDS(CID4, file= file_out)
}
```


## ML: Searchung the best SEM models

### SEM set-up

```{r}
lam1=0.3
lam2=0.1
q=0.2
mu=0.01

Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- pars.hc['lam002'] <- pars.hc['lam022'] <- lam1/3
pars.hc['lam200'] <- pars.hc['lam202'] <- pars.hc['lam222'] <- lam1/3

pars.hc['lam111'] <- pars.hc['lam113'] <- pars.hc['lam133'] <- lam1/3
pars.hc['lam311'] <- pars.hc['lam313'] <- pars.hc['lam333'] <- lam1/3

pars.hc['lam444'] <- lam2
pars.hc['lam555'] <- lam2
pars.hc['lam666'] <- lam2
pars.hc['lam777'] <- lam2
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-mu




```

parameter names

```{r}
# a0 a1 b0 b1 c0 c1 d0 d1
ar=names2array(args$arrays, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))
ar.re=reoder_lambdas(ar, c(1,3,5,7, 2,4,6,8))
#ar.re$lam.tensor


element <- c(
'lam000', 
'lam002', 
'lam022', 
'lam111',
'lam113',
'lam133',  
'lam200',
'lam202',
'lam222', 
'lam311',
'lam313',  
'lam333', 
'lam444',
'lam555',
'lam666',  
'lam777'
)


find_elements(ar.re$lam.tensor, element)
```

### SEM8:  100 tips

#### run

This algorithm selects the optimal rate matrixby iteratively testing 36 different models for each simulation trial. It works by dividing Q into a 4x4 block matrix, based on the states: a0b0; c0d0; a1b1; c1d1. For each block in a given row, the algorithm tests three different parameterizations and selects the best one before moving to the next block. Once a row is completed, the algorithm proceeds to the next row. Since there are four block rows in total, this results in testing 36 different models.

```{r, eval=FALSE}

# Define blocks
blocks <- list(
  matrix(c(1, 1, 1, 0), 2, byrow = TRUE),
  matrix(c(1, 1, 0, 0), 2, byrow = TRUE),
  matrix(c(1, 0, 0, 0), 2, byrow = TRUE)
)


# Define index
index <- list(
  matrix(c(1:2, 3:4,  1:2, 5:6,  1:2, 7:8), 3, 4, byrow = TRUE),
  matrix(c(3:4, 1:2,  3:4, 7:8,  3:4, 5:6), 3, 4, byrow = TRUE),
  matrix(c(5:6, 1:2,  5:6, 7:8,  5:6, 3:4), 3, 4, byrow = TRUE),
  matrix(c(7:8, 3:4,  7:8, 5:6,  7:8, 1:2), 3, 4, byrow = TRUE)
)

# Read phylogeny from file
base <- "phy_CID4-100tips-100tr"
file <- file.path("R/data/sim", paste0(base, '.RDS'))
print(paste0('Reading file: ', file))
phy <- readRDS(file = file)

# results
best_Qs <- list()
best_results <- list()
#SIM=1
for (SIM in 1:NSIM){
#for (SIM in 1:2){
  tree <- phy[[SIM]]
  states <- mapvalues(tree$tip.state, from = c("0", "1", "2", "3"), to = c(0, 1, 0, 1))
  root <- rep(1/8, 8)
  
  # Initialize result list
  # Q_ehe8_C.r
  Qtmp <- Q_ehe8_C.t
  result <- list()
  
  # Iterate over index
  #i=1
  for (i in seq_along(index)) {
    index.mat <- index[[i]]
    
    # Iterate over Q rows
    # j=1
    for (j in 1:nrow(index.mat)) {
      cells <- index.mat[j, ]
      # Update Qnew with block
      Qnew <- Qtmp
      # Iterate over precooked blocks
      # bi=1
      for (bi in 1:3) {
        print(paste('Working on tree: ',  SIM))
        print(paste('matrix: ',  i, j, bi))
        
        # Make new Qnew
        Qnew[cells[1:2], cells[3:4]] <- blocks[[bi]]
        
        # Make pars for inference
        v=c(1,5,2,6, 3,7,4,8)
        Qnew.r <- Qnew[v,v]
        qs <- extract_off_diagonal(Qnew.r/10)
        qsl=length(qs)
        pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
        zero.constr <- formulas_zero_pars(pars.hc)
        f.qs <- assign_classes_pairwise(Qnew.r/10, args$arrays$Q)
        f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
        f.lams <- c(lam002 ~ q01, lam022 ~ q01,
                    lam200 ~ q01, lam202 ~ q01, lam222 ~ q01,
                    lam111 ~ q01, lam113 ~ q01, lam133 ~ q01, 
                    lam311 ~ q01, lam313 ~ q01, lam333 ~ q01,
                    lam444 ~ q01, lam555 ~ q01, lam666 ~ q01, lam777 ~ q01,
                    lam000 ~ q01
        )
        f.list<- c(zero.constr,  f.qs, f.mu, f.lams)
        
        # Perform ML inference
        lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
        lik.const <- constrain(lik.c, formulae = f.list)
        arg.const <- argnames(lik.const)
        starting.point <- pars.hc[arg.const]
        mle <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
        
        # Store result
        b_ind <- (j - 1) * 3 + bi
        result[[b_ind]] <- mle
        result[[b_ind]]$sim_index <- c(j, bi)
        print(result[[b_ind]]$lnLik)
      }
    }
    # Select best and update Qtmp
    ln <- get_item(result, 'lnLik')
    #ln
    maxln <- which.max(ln)
    
    best <- result[[maxln]]
    sim_index <- best$sim_index
    cells <- index.mat[sim_index[1], ]
    Qtmp[cells[1:2], cells[3:4]] <- blocks[[sim_index[2]]]
  }
  best_results[[SIM]] <- best
  best_Qs[[SIM]] <- Qtmp
}



```


#### save
```{r}
# file_out=paste0('R/data/out/SEM/best_Ln-SEM8_', base, '.RDS')
# saveRDS(best_results, file= file_out)
# 
# file_out=paste0('R/data/out/SEM/best_Qs-SEM8_', base, '.RDS')
# saveRDS(best_Qs, file= file_out)
```


#### check pars

```{r}
# Plot the graph
ar <- pars_to_arrays(mle$par.full, 8)
ar <- names2array(ar, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))
Qpl <- ar$Q
g <- create_graph_from_matrix(Qpl)
unidirectional_edges <- is_unidirectional(g)
plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))

ar %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```

#### results

```{r}
# get_item(result, 'lnLik')
# # -318.1521 -318.1543 -318.6458 -318.1537 -318.1543 -318.1841
# max(get_item(result, 'lnLik'))
# # -318.1521

cid.lik <- c()
cid.lik$t100=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik')

dln=cid.lik$t100 - get_item(best_results, 'lnLik')
#best_Qs
hist(dln)
min(dln)
max(dln)

aic=get_item(best_results, 'lnLik') %>% get_aic(., 2)
aic_cid=get_aic(cid.lik$t100, 4)
daic=aic_cid  -  aic
hist(daic)
max(daic)
```

### SEM8: 500 tips

#### run

```{r, eval=FALSE}

# Read phylogeny from file
base <- "phy_CID4-500tips-100tr"
file <- file.path("R/data", paste0(base, '.RDS'))
print(paste0('Reading file: ', file))
phy <- readRDS(file = file)

# results
best_Qs <- list()
best_results <- list()
#SIM=1
for (SIM in 1:NSIM){
#for (SIM in 1:2){
  tree <- phy[[SIM]]
  states <- mapvalues(tree$tip.state, from = c("0", "1", "2", "3"), to = c(0, 1, 0, 1))
  root <- rep(1/8, 8)
  
  # Initialize result list
  # Q_ehe8_C.r
  Qtmp <- Q_ehe8_C.t
  result <- list()
  
  # Iterate over index
  i=1
  for (i in seq_along(index)) {
    index.mat <- index[[i]]
    
    # Iterate over Q rows
    for (j in 1:nrow(index.mat)) {
      cells <- index.mat[j, ]
      # Update Qnew with block
      Qnew <- Qtmp
      # Iterate over precooked blocks
      for (bi in 1:3) {
        print(paste('Working on tree: ',  SIM))
        print(paste('matrix: ',  i, j, bi))
        
        # Make new Qnew
        Qnew[cells[1:2], cells[3:4]] <- blocks[[bi]]
        
        # Make pars for inference
        v=c(1,5,2,6, 3,7,4,8)
        Qnew.r <- Qnew[v,v]
        qs <- extract_off_diagonal(Qnew.r/10)
        qsl=length(qs)
        pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
        zero.constr <- formulas_zero_pars(pars.hc)
        f.qs <- assign_classes_pairwise(Qnew.r/10, args$arrays$Q)
        f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
        f.lams <- c(lam002 ~ q01, lam022 ~ q01,
                    lam200 ~ q01, lam202 ~ q01, lam222 ~ q01,
                    lam111 ~ q01, lam113 ~ q01, lam133 ~ q01, 
                    lam311 ~ q01, lam313 ~ q01, lam333 ~ q01,
                    lam444 ~ q01, lam555 ~ q01, lam666 ~ q01, lam777 ~ q01,
                    lam000 ~ q01
        )
        f.list<- c(zero.constr,  f.qs, f.mu, f.lams)
        
        # Perform ML inference
        lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
        lik.const <- constrain(lik.c, formulae = f.list)
        arg.const <- argnames(lik.const)
        starting.point <- pars.hc[arg.const]
        mle <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
        
        # Store result
        b_ind <- (j - 1) * 3 + bi
        result[[b_ind]] <- mle
        result[[b_ind]]$sim_index <- c(j, bi)
        print(result[[b_ind]]$lnLik)
      }
    }
    # Select best and update Qtmp
    ln <- get_item(result, 'lnLik')
    #ln
    maxln <- which.max(ln)
    
    best <- result[[maxln]]
    sim_index <- best$sim_index
    cells <- index.mat[sim_index[1], ]
    Qtmp[cells[1:2], cells[3:4]] <- blocks[[sim_index[2]]]
  }
  best_results[[SIM]] <- best
  best_Qs[[SIM]] <- Qtmp
}



```

#### save

```{r}
# file_out=paste0('R/data/non-congruent/best_results-EHE8_', base, '.RDS')
# saveRDS(best_results, file= file_out)
# # 
# file_out=paste0('R/data/non-congruent/best_Qs-EHE8_', base, '.RDS')
# saveRDS(best_Qs, file= file_out)
```


#### res

```{r}
# get_item(result, 'lnLik')
# # -318.1521 -318.1543 -318.6458 -318.1537 -318.1543 -318.1841
# max(get_item(result, 'lnLik'))
# # -318.1521

dln=cid.lik$t500-get_item(best_results, 'lnLik')
#best_Qs
hist(dln)
min(dln)
max(dln)

aic=get_item(best_results, 'lnLik') %>% get_aic(., 2)
daic=cid$t500  -  aic
hist(daic)
max(daic)
```




### -- 1000 tips

#### run

```{r, eval=FALSE}

# Read phylogeny from file
base <- "phy_CID4-1000tips-100tr"
file <- file.path("R/data", paste0(base, '.RDS'))
print(paste0('Reading file: ', file))
phy <- readRDS(file = file)
#plot(phy[[1]])

# files to save
# file_out1=paste0('R/data/non-congruent/best_results-EHE8_', base, '.RDS')
# file_out2=paste0('R/data/non-congruent/best_Qs-EHE8_', base, '.RDS')
file_out1=paste0('R/data/non-congruent/55-best_results-EHE8_', base, '.RDS')
file_out2=paste0('R/data/non-congruent/55-best_Qs-EHE8_', base, '.RDS')



# results
best_Qs <- list()
best_results <- list()
#SIM=1
#for (SIM in 1:NSIM){
for (SIM in 55:NSIM){
  tree <- phy[[SIM]]
  states <- mapvalues(tree$tip.state, from = c("0", "1", "2", "3"), to = c(0, 1, 0, 1))
  root <- rep(1/8, 8)
  
  # Initialize result list
  # Q_ehe8_C.r
  Qtmp <- Q_ehe8_C.t
  result <- list()
  
  # Iterate over index
  i=1
  for (i in seq_along(index)) {
    index.mat <- index[[i]]
    
    # Iterate over Q rows
    for (j in 1:nrow(index.mat)) {
      cells <- index.mat[j, ]
      # Update Qnew with block
      Qnew <- Qtmp
      # Iterate over precooked blocks
      for (bi in 1:3) {
        print(paste('Working on tree: ',  SIM))
        print(paste('matrix: ',  i, j, bi))
        
        # Make new Qnew
        Qnew[cells[1:2], cells[3:4]] <- blocks[[bi]]
        
        # Make pars for inference
        v=c(1,5,2,6, 3,7,4,8)
        Qnew.r <- Qnew[v,v]
        qs <- extract_off_diagonal(Qnew.r/10)
        qsl=length(qs)
        pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
        zero.constr <- formulas_zero_pars(pars.hc)
        f.qs <- assign_classes_pairwise(Qnew.r/10, args$arrays$Q)
        f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
        f.lams <- c(lam002 ~ q01, lam022 ~ q01,
                    lam200 ~ q01, lam202 ~ q01, lam222 ~ q01,
                    lam111 ~ q01, lam113 ~ q01, lam133 ~ q01, 
                    lam311 ~ q01, lam313 ~ q01, lam333 ~ q01,
                    lam444 ~ q01, lam555 ~ q01, lam666 ~ q01, lam777 ~ q01,
                    lam000 ~ q01
        )
        f.list<- c(zero.constr,  f.qs, f.mu, f.lams)
        
        # Perform ML inference
        lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
        lik.const <- constrain(lik.c, formulae = f.list)
        arg.const <- argnames(lik.const)
        starting.point <- pars.hc[arg.const]
        mle <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
        
        # Store result
        b_ind <- (j - 1) * 3 + bi
        result[[b_ind]] <- mle
        result[[b_ind]]$sim_index <- c(j, bi)
        print(result[[b_ind]]$lnLik)
      }
    }
    # Select best and update Qtmp
    ln <- get_item(result, 'lnLik')
    #ln
    maxln <- which.max(ln)
    
    best <- result[[maxln]]
    sim_index <- best$sim_index
    cells <- index.mat[sim_index[1], ]
    Qtmp[cells[1:2], cells[3:4]] <- blocks[[sim_index[2]]]
  }
  best_results[[SIM]] <- best
  best_Qs[[SIM]] <- Qtmp
  
  # save
  saveRDS(best_results, file= file_out1)
  saveRDS(best_Qs, file= file_out2)
}



```

```{r}
#best_Qs
```

```{r}
# res1 <- readRDS('R/data/non-congruent/best_results-EHE8_phy_CID4-1000tips-100tr.RDS')
# length(res1)
# sapply(res1, length)
# res2 <- readRDS('R/data/non-congruent/55-best_results-EHE8_phy_CID4-1000tips-100tr.RDS')
# sapply(res2, length)
# length(res2)
# 
# best_results <- res2
# best_results[1:54] <- res1
# sapply(best_results, length)
#saveRDS(best_results, file= 'best_results-EHE8_phy_CID4-1000tips-100tr.RDS')

b <- readRDS('best_results-EHE8_phy_CID4-1000tips-100tr.RDS')
length(b)
sapply(b, length)
get_item(b, 'lnlik')

# q1 <- readRDS('R/data/non-congruent/best_Qs-EHE8_phy_CID4-1000tips-100tr.RDS')
# length(q1)
# sapply(q1, length)
# q2 <- readRDS('R/data/non-congruent/55-best_Qs-EHE8_phy_CID4-1000tips-100tr.RDS')
# sapply(q2, length)
# length(q2)
# 
# best_Qs <- q2
# best_Qs[1:54] <- q1
# sapply(best_Qs, length)
# saveRDS(best_Qs, file= 'best_best_Qs-EHE8_phy_CID4-1000tips-100tr.RDS')

```



#### res

```{r}
# get_item(result, 'lnLik')
# # -318.1521 -318.1543 -318.6458 -318.1537 -318.1543 -318.1841
# max(get_item(result, 'lnLik'))
# # -318.1521

dln=cid.lik$t1000-get_item(best_results, 'lnLik')
#best_Qs
hist(dln)
min(dln)
max(dln)

aic=get_item(best_results, 'lnLik') %>% get_aic(., 2)
daic=cid$t1000  -  aic
hist(daic)
max(daic)
```




## Results


```{r}

# cid <- tibble(
# t100=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4),
# t500=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4),
# t1000=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4)
# )
# 
# ehe <- tibble(
# t100=get_item(readRDS(file='R/data/lik-EHE8_C-phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2),
# t500=get_item(readRDS(file='R/data/lik-EHE8_C-phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2),
# t1000=get_item(readRDS(file='R/data/lik-EHE8_C-phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2)
# )
# 
# 
# cid.lik <- tibble(
# t100=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik'),
# t500=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik') ,
# t1000=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik')
# )
# 
# ehe.lik <- tibble(
# t100=get_item(readRDS(file='R/data/lik-EHE8_C-phy_CID4-100tips-100tr.RDS'), 'lnLik'),
# t500=get_item(readRDS(file='R/data/lik-EHE8_C-phy_CID4-500tips-100tr.RDS'), 'lnLik'),
# t1000=get_item(readRDS(file='R/data/lik-EHE8_C-phy_CID4-1000tips-100tr.RDS'), 'lnLik')
# )

```


```{r}


cid <- tibble(
t100=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4),
t500=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4),
t1000=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4)
)

cid.lik <- tibble(
t100=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik'),
t500=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik') ,
t1000=get_item(readRDS(file='R/data/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik')
)

noncong.aic <- tibble(
t100=get_item(readRDS( file='R/data/non-congruent/best_results-EHE8_phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2),
t500=get_item(readRDS( file='R/data/non-congruent/best_results-EHE8_phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2),
t1000=get_item(readRDS(file='R/data/non-congruent/best_results-EHE8_phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2)
)


noncong.lik <- tibble(
t100=get_item(readRDS( file='R/data/non-congruent/best_results-EHE8_phy_CID4-100tips-100tr.RDS'), 'lnLik'),
t500=get_item(readRDS( file='R/data/non-congruent/best_results-EHE8_phy_CID4-500tips-100tr.RDS'), 'lnLik'),
t1000=get_item(readRDS(file='R/data/non-congruent/best_results-EHE8_phy_CID4-1000tips-100tr.RDS'), 'lnLik')
)

```

MLE of CID is constantly better than MLE of the semi-congruent EHE8-C.

```{r}
del.lik.nc <- cid.lik - noncong.lik

# The positive values indicate that CID's MLE is better, negative that noncong is better
# There are only positive values
apply(del.lik.nc, 2, min)

#par(mfrow = c(1, 3))
hist(del.lik.nc$t100, breaks = 30)
hist(del.lik.nc$t500, breaks = 30)
hist(del.lik.nc$t1000, breaks = 30)

hist(del.lik.nc$t100*2, breaks = 30)
hist(del.lik.nc$t500*2, breaks = 30)
hist(del.lik.nc$t1000*2, breaks = 30)

```


```{r}

library(ggplot2)
library(dplyr)
library(tidyr)  # Load tidyr package

del.nc <- cid-noncong.aic
apply(del.nc, 2, max)

# Calculate ECDF for each column
ecdf_del_nc <- del.nc %>%
  summarise(
    t100_ecdf = ecdf(t100),
    t500_ecdf = ecdf(t500),
    t1000_ecdf = ecdf(t1000)
  )

# Create a data frame for plotting
ecdf_data_nc <- data.frame(
  Value = c(del.nc$t100, del.nc$t500, del.nc$t1000),
  Group = rep(c("100 tips", "500 tips", "1000 tips"), each = nrow(del.nc))
)


#hist(del.nc$t1000, breaks=30)
```

Proportion of trials with dAIC > 2.

```{r}

1-ecdf_del_nc$t100_ecdf(2)
1-ecdf_del_nc$t500_ecdf(2)
1-ecdf_del_nc$t1000_ecdf(2)

```

Plot NOT all

```{r}


# Define linetypes for each group
linetypes <- c("100 tips" = "dotted", "500 tips" = "dashed", "1000 tips" = "solid")

# Plot the ECDFs for each group with varying linetypes
cdf_plot <- ggplot() +
  stat_ecdf(data = ecdf_data_nc, aes(x = Value, linetype = Group), geom = "step", pad = F, size = .4) +
  #stat_ecdf(data = d_semicong, aes(x = Value, linetype = Group, color = type), geom = "step", pad = F, size = .4) +
  labs(x = expression(Delta~AIC), y = "CDF") +
  theme_minimal()+
  scale_x_continuous(
    limits = c(-4, 12),
    #breaks = seq(-4, 12, by = 1)
    breaks = seq(-4, 12, by = 2)
  ) +
  scale_linetype_manual(values = linetypes) + # Apply manual linetype scale
  theme(
    panel.grid.minor.x = element_blank(),  # Remove minor grid lines on the x-axis
    panel.grid.minor.y = element_blank()
  ) +
  #guides(color = guide_legend(title = NULL)) +
  guides(color = "none") +
  geom_vline(xintercept = 2, linetype = "dashed", color = "black", size = 0.5) +
  theme(
    #plot.background = element_rect(fill = "white"),
    #panel.background = element_rect(fill = "white"),
    legend.box.background = element_rect(fill = "white"),
    legend.text = element_text(size = 6),
    legend.key.size = unit(1.1, "lines"),
    legend.title = element_text(size = 0),
    legend.position = c(0.15, 0.70)
)

cdf_plot

# Save the plot with a specific width in centimeters
# ggsave("Figs/cdf-nc-c.eps", plot = cdf_plot , width = 8.7, height = 5, units = "cm")
# ggsave("Figs/cdf-nc-c.png", plot = cdf_plot , width = 8.7, height = 5, units = "cm")

```



### Unique Q matrices

```{r}
# 100 tips
Qs <- readRDS( file='R/data/non-congruent/best_Qs-EHE8_phy_CID4-100tips-100tr.RDS')
vector_list <-lapply(Qs, function(x) c(x))
vector_matrix <- do.call(rbind, vector_list)
unique_vectors <- unique(vector_matrix)
nrow(unique_vectors)
# [1] 95

# 500 tips
Qs <- readRDS( file='R/data/non-congruent/best_Qs-EHE8_phy_CID4-500tips-100tr.RDS')
vector_list <-lapply(Qs, function(x) c(x))
vector_matrix <- do.call(rbind, vector_list)
unique_vectors <- unique(vector_matrix)
nrow(unique_vectors)
# [1] 86

# 1000 tips
Qs <- readRDS( file='R/data/non-congruent/best_Qs-EHE8_phy_CID4-1000tips-100tr.RDS')
vector_list <-lapply(Qs, function(x) c(x))
vector_matrix <- do.call(rbind, vector_list)
unique_vectors <- unique(vector_matrix)
nrow(unique_vectors)
# [1] 70
```


### Plot matrix

```{r}
del.nc
del.nc[2,]
Qs <- readRDS( file='R/data/non-congruent/best_Qs-EHE8_phy_CID4-1000tips-100tr.RDS')
Qfocal <- Qs[[2]]

Qfocal
# Plot the graph
g <- create_graph_from_matrix(Qfocal)
unidirectional_edges <- is_unidirectional(g)
plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))
```

