---
#title: "NON-Semi-Congruent"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
params:
  output_dir: "/output"
cache: TRUE
---

# SEM: semicongruent models


In this example, data are simulated under the CID4 model for three different scenarios: 100, 500, and 1000 tips. We conduct ML inference using the original CID models and a set of SEM models which are optimized iteratively over 36 steps to select the best SEM model by testing various Q matrices.

This algorithm  works by dividing Q into a 4x4 block matrix, based on the states: a0b0; c0d0; a1b1; c1d1. For each block in a given row, the algorithm tests three different parameterizations and selects the best one before moving to the next block. Once a row is completed, the algorithm proceeds to the next row. Since there are four block rows in total, this results in testing 36 different models.

The tested SEM. models constantly show better AIC and BIC compared to the original CID model.


```{r,  include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here())
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
source('R/utils/dependencies.R')
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```

## Setting-up Q matrices for inference

### CID4

```{r}
# trait
Q_t <- initQ(c(0,1), c(2, 2))
# diversification regime
Q_r2 <- initQ(c('A','B'), c(2,2))
# speciation rates for tracking the order
La4 <- diag(c(-3,-1),2)
La4 = La4 %x% diag(1,2)
# order according to regimes
Q_cid4.r <- amaSMM(Q_r2, Q_t)

# order according to trait
v=c(1,3, 2,4)
Q_cid4.t <- Q_cid4.r[v,v]

print(Q_cid4.r)
print(La4)
print(Q_cid4.t)
print(La4[v,v])


```

### SEM8

Similar to EHE model.

```{r}

# Create the matrix as a data frame
Q_ehe8_C.t <- data.frame(
  a0 = c(-4, 0, 1, 1, 1, 1, 0, 0),
  b0 = c(0, -4, 1, 1, 1, 1, 0, 0),
  c0 = c(1, 1, -4, 0, 0, 0, 1, 1),
  d0 = c(1, 1, 0, -4, 0, 0, 1, 1),
  a1 = c(1, 1, 0, 0, -4, 0, 1, 1),
  b1 = c(1, 1, 0, 0, 0, -4, 1, 1),
  c1 = c(0, 0, 1, 1, 1, 1, -4, 0),
  d1 = c(0, 0, 1, 1, 1, 1, 0, -4)
)

# Set row names and column names
rownames(Q_ehe8_C.t) <- colnames(Q_ehe8_C.t)
Q_ehe8_C.t <-as.matrix(Q_ehe8_C.t)


v=c(1,5, 2,6, 3,7, 4,8)
Q_ehe8_C.r <- Q_ehe8_C.t[v,v]

print(Q_ehe8_C.r)
#print(La8)
print(Q_ehe8_C.t)
#print(La8[v,v])

```


## Read in data

```{r}

NSIM=100
files_base <- c('phy_CID4-100tips-100tr', 'phy_CID4-500tips-100tr', 'phy_CID4-1000tips-100tr')

```

## ML: CID4 (4 pars)

```{r}


Args <- list(
  Nstates = 4L,
  y = list(
    c(0,0,0,0, 1,0,1,0),
    c(0,0,0,0, 0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(4)
#args$arrays
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.05
pars.hc['lam333'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <-0.1
qs <- extract_off_diagonal(Q_cid4.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,4)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid4.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam333 ~ lam222)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r, eval=FALSE}
#base="phy_CID4-100tips-100tr"
for (base in files_base[1:3]){
  file=file.path("R/data/sim", paste0(base, '.RDS'))
  print(paste0('Reading file: ',file))
  phy <- readRDS(file=file)

  CID4 <- list()
  #i=1
  #which(is.na(phy))
  for (i in 1:NSIM){

    print(paste0('Working on: ', i))
    tree <- phy[[i]]
    states<- tree$tip.state
    states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
    root <- c(1/4, 1/4, 1/4, 1/4)

    lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
    lik.const <- constrain(lik.c, formulae = f.list)
    arg.const <- argnames(lik.const)
    starting.point <- pars.hc[arg.const]
    CID4[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
    #CID4[[1]]$lnLik

  }
  file_out=paste0('R/data/out/lik-CID4-', base, '.RDS')
  saveRDS(CID4, file= file_out)
}
```


## ML: Searchung the best SEM models

### SEM set-up

```{r}
lam1=0.3
lam2=0.1
q=0.2
mu=0.01

Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- pars.hc['lam002'] <- pars.hc['lam022'] <- lam1/3
pars.hc['lam200'] <- pars.hc['lam202'] <- pars.hc['lam222'] <- lam1/3

pars.hc['lam111'] <- pars.hc['lam113'] <- pars.hc['lam133'] <- lam1/3
pars.hc['lam311'] <- pars.hc['lam313'] <- pars.hc['lam333'] <- lam1/3

pars.hc['lam444'] <- lam2
pars.hc['lam555'] <- lam2
#pars.hc['lam666'] <- lam2
pars.hc['lam646'] <- lam2
pars.hc['lam777'] <- lam2
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-mu


# Define blocks
blocks <- list(
  matrix(c(1, 1, 1, 0), 2, byrow = TRUE),
  matrix(c(1, 1, 0, 0), 2, byrow = TRUE),
  matrix(c(1, 0, 0, 0), 2, byrow = TRUE)
)


# Define index
index <- list(
  matrix(c(1:2, 3:4,  1:2, 5:6,  1:2, 7:8), 3, 4, byrow = TRUE),
  matrix(c(3:4, 1:2,  3:4, 7:8,  3:4, 5:6), 3, 4, byrow = TRUE),
  matrix(c(5:6, 1:2,  5:6, 7:8,  5:6, 3:4), 3, 4, byrow = TRUE),
  matrix(c(7:8, 3:4,  7:8, 5:6,  7:8, 1:2), 3, 4, byrow = TRUE)
)

```

parameter names

```{r}
# a0 a1 b0 b1 c0 c1 d0 d1
ar=names2array(args$arrays, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))
ar.re=reoder_lambdas(ar, c(1,3,5,7, 2,4,6,8))
#ar.re$lam.tensor


element <- c(
'lam000', 
'lam002', 
'lam022', 
'lam111',
'lam113',
'lam133',  
'lam200',
'lam202',
'lam222', 
'lam311',
'lam313',  
'lam333', 
'lam444',
'lam555',
'lam646',  
'lam777'
)


find_elements(ar.re$lam.tensor, element)
```

### SEM8:  100 tips

#### run



```{r, eval=FALSE}

# Read phylogeny from file
base <- "phy_CID4-100tips-100tr"
file <- file.path("R/data/sim", paste0(base, '.RDS'))
print(paste0('Reading file: ', file))
phy <- readRDS(file = file)

# results
best_Qs <- list()
best_results <- list()
#SIM=1
for (SIM in 1:NSIM){
#for (SIM in 1:2){
  tree <- phy[[SIM]]
  states <- mapvalues(tree$tip.state, from = c("0", "1", "2", "3"), to = c(0, 1, 0, 1))
  root <- rep(1/8, 8)
  
  # Initialize result list
  # Q_ehe8_C.r
  Qtmp <- Q_ehe8_C.t
  result <- list()
  
  # Iterate over index
  #i=1
  for (i in seq_along(index)) {
    index.mat <- index[[i]]
    
    # Iterate over Q rows
    # j=1
    for (j in 1:nrow(index.mat)) {
      cells <- index.mat[j, ]
      # Update Qnew with block
      Qnew <- Qtmp
      # Iterate over precooked blocks
      # bi=1
      for (bi in 1:3) {
        print(paste('Working on tree: ',  SIM))
        print(paste('matrix: ',  i, j, bi))
        
        # Make new Qnew
        Qnew[cells[1:2], cells[3:4]] <- blocks[[bi]]
        
        # Make pars for inference
        v=c(1,5,2,6, 3,7,4,8)
        Qnew.r <- Qnew[v,v]
        qs <- extract_off_diagonal(Qnew.r/10)
        qsl=length(qs)
        pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
        zero.constr <- formulas_zero_pars(pars.hc)
        f.qs <- assign_classes_pairwise(Qnew.r/10, args$arrays$Q)
        f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
        f.lams <- c(lam002 ~ q01, lam022 ~ q01,
                    lam200 ~ q01, lam202 ~ q01, lam222 ~ q01,
                    lam111 ~ q01, lam113 ~ q01, lam133 ~ q01, 
                    lam311 ~ q01, lam313 ~ q01, lam333 ~ q01,
                    lam444 ~ q01, lam555 ~ q01, lam646 ~ q01, lam777 ~ q01,
                    lam000 ~ q01
        )
        f.list<- c(zero.constr,  f.qs, f.mu, f.lams)
        
        # Perform ML inference
        lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
        lik.const <- constrain(lik.c, formulae = f.list)
        arg.const <- argnames(lik.const)
        starting.point <- pars.hc[arg.const]
        mle <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
        
        # Store result
        b_ind <- (j - 1) * 3 + bi
        result[[b_ind]] <- mle
        result[[b_ind]]$sim_index <- c(j, bi)
        print(result[[b_ind]]$lnLik)
      }
    }
    # Select best and update Qtmp
    ln <- get_item(result, 'lnLik')
    #ln
    maxln <- which.max(ln)
    
    best <- result[[maxln]]
    sim_index <- best$sim_index
    cells <- index.mat[sim_index[1], ]
    Qtmp[cells[1:2], cells[3:4]] <- blocks[[sim_index[2]]]
  }
  best_results[[SIM]] <- best
  best_Qs[[SIM]] <- Qtmp
}



```


#### save
```{r}
# file_out=paste0('R/data/out/SEM/best_Ln-SEM8_', base, '.RDS')
# saveRDS(best_results, file= file_out)
# 
# file_out=paste0('R/data/out/SEM/best_Qs-SEM8_', base, '.RDS')
# saveRDS(best_Qs, file= file_out)
```


#### check pars

```{r, eval=FALSE}
# Plot the graph
ar <- pars_to_arrays(mle$par.full, 8)
ar <- names2array(ar, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))
Qpl <- ar$Q
g <- create_graph_from_matrix(Qpl)
unidirectional_edges <- is_unidirectional(g)
plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))

ar %>% reoder_lambdas(., c(1,3,5,7, 2,4,6,8))
```


### SEM8: 500 tips

#### run

```{r, eval=FALSE}

# Read phylogeny from file
base <- "phy_CID4-500tips-100tr"
file <- file.path("R/data/sim", paste0(base, '.RDS'))
print(paste0('Reading file: ', file))
phy <- readRDS(file = file)

# results
best_Qs <- list()
best_results <- list()
#SIM=1
for (SIM in 1:NSIM){
#for (SIM in 1:2){
  tree <- phy[[SIM]]
  states <- mapvalues(tree$tip.state, from = c("0", "1", "2", "3"), to = c(0, 1, 0, 1))
  root <- rep(1/8, 8)
  
  # Initialize result list
  # Q_ehe8_C.r
  Qtmp <- Q_ehe8_C.t
  result <- list()
  
  # Iterate over index
  i=1
  for (i in seq_along(index)) {
    index.mat <- index[[i]]
    
    # Iterate over Q rows
    for (j in 1:nrow(index.mat)) {
      cells <- index.mat[j, ]
      # Update Qnew with block
      Qnew <- Qtmp
      # Iterate over precooked blocks
      for (bi in 1:3) {
        print(paste('Working on tree: ',  SIM))
        print(paste('matrix: ',  i, j, bi))
        
        # Make new Qnew
        Qnew[cells[1:2], cells[3:4]] <- blocks[[bi]]
        
        # Make pars for inference
        v=c(1,5,2,6, 3,7,4,8)
        Qnew.r <- Qnew[v,v]
        qs <- extract_off_diagonal(Qnew.r/10)
        qsl=length(qs)
        pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
        zero.constr <- formulas_zero_pars(pars.hc)
        f.qs <- assign_classes_pairwise(Qnew.r/10, args$arrays$Q)
        f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
        f.lams <- c(lam002 ~ q01, lam022 ~ q01,
                    lam200 ~ q01, lam202 ~ q01, lam222 ~ q01,
                    lam111 ~ q01, lam113 ~ q01, lam133 ~ q01, 
                    lam311 ~ q01, lam313 ~ q01, lam333 ~ q01,
                    lam444 ~ q01, lam555 ~ q01, lam646 ~ q01, lam777 ~ q01,
                    lam000 ~ q01
        )
        f.list<- c(zero.constr,  f.qs, f.mu, f.lams)
        
        # Perform ML inference
        lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
        lik.const <- constrain(lik.c, formulae = f.list)
        arg.const <- argnames(lik.const)
        starting.point <- pars.hc[arg.const]
        mle <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
        
        # Store result
        b_ind <- (j - 1) * 3 + bi
        result[[b_ind]] <- mle
        result[[b_ind]]$sim_index <- c(j, bi)
        print(result[[b_ind]]$lnLik)
      }
    }
    # Select best and update Qtmp
    ln <- get_item(result, 'lnLik')
    #ln
    maxln <- which.max(ln)
    
    best <- result[[maxln]]
    sim_index <- best$sim_index
    cells <- index.mat[sim_index[1], ]
    Qtmp[cells[1:2], cells[3:4]] <- blocks[[sim_index[2]]]
  }
  best_results[[SIM]] <- best
  best_Qs[[SIM]] <- Qtmp
}



```

#### save

```{r}

# file_out=paste0('R/data/out/SEM/best_Ln-SEM8_', base, '.RDS')
# saveRDS(best_results, file= file_out)
# 
# file_out=paste0('R/data/out/SEM/best_Qs-SEM8_', base, '.RDS')
# saveRDS(best_Qs, file= file_out)

```


#### results

```{r, eval=FALSE}
cid.aic=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4)
sem.aic=get_item(readRDS(file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2)
daic=cid.aic-sem.aic
hist(daic)
```




### SEM8: 1000 tips

#### run

```{r, eval=FALSE}

# Read phylogeny from file
base <- "phy_CID4-1000tips-100tr"
file <- file.path("R/data/sim", paste0(base, '.RDS'))
print(paste0('Reading file: ', file))
phy <- readRDS(file = file)
#plot(phy[[1]])

# files to save
file_out1=paste0('R/data/out/SEM/best_Ln-SEM8_', base, '.RDS')
file_out2=paste0('R/data/out/SEM/best_Qs-SEM8_', base, '.RDS')


# results
best_Qs <- list()
best_results <- list()
#SIM=1
for (SIM in 1:NSIM){
#for (SIM in 1:50){
#for (SIM in 51:NSIM){
  tree <- phy[[SIM]]
  states <- mapvalues(tree$tip.state, from = c("0", "1", "2", "3"), to = c(0, 1, 0, 1))
  root <- rep(1/8, 8)
  
  # Initialize result list
  # Q_ehe8_C.r
  Qtmp <- Q_ehe8_C.t
  result <- list()
  
  # Iterate over index
  i=1
  for (i in seq_along(index)) {
    index.mat <- index[[i]]
    
    # Iterate over Q rows
    for (j in 1:nrow(index.mat)) {
      cells <- index.mat[j, ]
      # Update Qnew with block
      Qnew <- Qtmp
      # Iterate over precooked blocks
      for (bi in 1:3) {
        print(paste('Working on tree: ',  SIM))
        print(paste('matrix: ',  i, j, bi))
        
        # Make new Qnew
        Qnew[cells[1:2], cells[3:4]] <- blocks[[bi]]
        
        # Make pars for inference
        v=c(1,5,2,6, 3,7,4,8)
        Qnew.r <- Qnew[v,v]
        qs <- extract_off_diagonal(Qnew.r/10)
        qsl=length(qs)
        pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
        zero.constr <- formulas_zero_pars(pars.hc)
        f.qs <- assign_classes_pairwise(Qnew.r/10, args$arrays$Q)
        f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
        f.lams <- c(lam002 ~ q01, lam022 ~ q01,
                    lam200 ~ q01, lam202 ~ q01, lam222 ~ q01,
                    lam111 ~ q01, lam113 ~ q01, lam133 ~ q01, 
                    lam311 ~ q01, lam313 ~ q01, lam333 ~ q01,
                    lam444 ~ q01, lam555 ~ q01, lam646 ~ q01, lam777 ~ q01,
                    lam000 ~ q01
        )
        f.list<- c(zero.constr,  f.qs, f.mu, f.lams)
        
        # Perform ML inference
        lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
        lik.const <- constrain(lik.c, formulae = f.list)
        arg.const <- argnames(lik.const)
        starting.point <- pars.hc[arg.const]
        mle <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
        
        # Store result
        b_ind <- (j - 1) * 3 + bi
        result[[b_ind]] <- mle
        result[[b_ind]]$sim_index <- c(j, bi)
        print(result[[b_ind]]$lnLik)
      }
    }
    # Select best and update Qtmp
    ln <- get_item(result, 'lnLik')
    #ln
    maxln <- which.max(ln)
    
    best <- result[[maxln]]
    sim_index <- best$sim_index
    cells <- index.mat[sim_index[1], ]
    Qtmp[cells[1:2], cells[3:4]] <- blocks[[sim_index[2]]]
  }
  best_results[[SIM]] <- best
  best_Qs[[SIM]] <- Qtmp
  
  # save
  saveRDS(best_results, file= file_out1)
  saveRDS(best_Qs, file= file_out2)
}



```


to speed up computations, I ran the above by parts, now combining the results

```{r, eval=FALSE}
# to speed computations, I ran them by parts
# combining the results for Ln

res1 <- readRDS("R/data/out/SEM/1-best_Ln-SEM8_phy_CID4-1000tips-100tr.RDS")
res2 <- readRDS("R/data/out/SEM/2-best_Ln-SEM8_phy_CID4-1000tips-100tr.RDS")
length(res1)
sapply(res1, length)
sapply(res2, length)
length(res2)

best_results <- res2
best_results[1:50] <- res1
sapply(best_results, length)
#saveRDS(best_results, file= "R/data/out/SEM/best_Ln-SEM8_phy_CID4-1000tips-100tr.RDS")



# combining the results for Qs
q1 <- readRDS("R/data/out/SEM/1-best_Qs-SEM8_phy_CID4-1000tips-100tr.RDS")
q2 <- readRDS("R/data/out/SEM/2-best_Qs-SEM8_phy_CID4-1000tips-100tr.RDS")
length(q1)
sapply(q1, length)
sapply(q2, length)
length(q2)

best_Qs <- q2
best_Qs[1:50] <- q1
sapply(best_Qs, length)
#saveRDS(best_Qs, file= "R/data/out/SEM/best_Qs-SEM8_phy_CID4-1000tips-100tr.RDS")

```



#### results

```{r}
cid.aic=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4)
sem.aic=get_item(readRDS(file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2)
daic=cid.aic-sem.aic
hist(daic)
```


## Results

```{r}
library(rphenoscate)
library(ggplot2)
library(dplyr)
library(tidyr)  # Load tidyr package
```

### Plot AIC results

```{r}

# read the results
cid.aic <- tibble(
  t100=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4),
  t500=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4),
  t1000=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4)
)

cid.lik <- tibble(
  t100=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik'),
  t500=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik'), 
  t1000=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik')
)

sem.aic <- tibble(
  t100=get_item(readRDS( file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2),
  t500=get_item(readRDS( file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2),
  t1000=get_item(readRDS(file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 2)
)


sem.lik <- tibble(
  t100=get_item(readRDS( file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-100tips-100tr.RDS'), 'lnLik'),
  t500=get_item(readRDS( file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-500tips-100tr.RDS'), 'lnLik'),
  t1000=get_item(readRDS(file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-1000tips-100tr.RDS'), 'lnLik')
)

```


MLE of SEM varies from being better than of the original CID to being worse.

```{r}
del.lik <- cid.lik - sem.lik

# The positive values indicate that CID's MLE is better, negative that SEM is better
apply(del.lik, 2, min)

hist(del.lik$t100, breaks = 30)
hist(del.lik$t500, breaks = 30)
hist(del.lik$t1000, breaks = 30)

# proportion of trials supportiing SEM
(which(del.lik$t100 < 0) %>% length)/100
(which(del.lik$t500 < 0) %>% length)/100
(which(del.lik$t1000 < 0) %>% length)/100

# [1] 0.58
# [1] 0.71
# [1] 0.56

```

Get dAIC

```{r}

# get max delta aic
del.aic <- cid.aic-sem.aic
apply(del.aic, 2, max)
apply(del.aic, 2, min)

# Calculate ECDF for each column
ecdf_del <- del.aic %>%
  summarise(
    t100_ecdf = ecdf(t100),
    t500_ecdf = ecdf(t500),
    t1000_ecdf = ecdf(t1000)
  )


# # Create a data frame for plotting
# ecdf_data <- data.frame(
#   Value = c(del.aic$t100, del.aic$t500),
#   Group = rep(c("100 tips", "500 tips"), each = nrow(del.aic))
# )

# Create a data frame for plotting
ecdf_data <- data.frame(
  Value = c(del.aic$t100, del.aic$t500, del.aic$t1000),
  Group = rep(c("100 tips", "500 tips", "1000 tips"), each = nrow(del.aic))
)


```

ECDF: Proportion of trials with dAIC > 2.

```{r}

1-ecdf_del$t100_ecdf(2)
1-ecdf_del$t500_ecdf(2)
1-ecdf_del$t1000_ecdf(2)

```

#### Plot AIC

```{r}


# Define linetypes for each group
linetypes <- c("100 tips" = "dotted", "500 tips" = "dashed", "1000 tips" = "solid")
colortypes <- c("100 tips" = "#619CFF", "500 tips" = "#00BA38", "1000 tips" = "#F8766D")
#linetypes <- c("100 tips" = "dotted", "500 tips" = "dashed")

# Plot the ECDFs for each group with varying linetypes
cdf_plot <- ggplot() +
  stat_ecdf(data = ecdf_data, aes(x = Value, linetype = Group, color=Group), geom = "step", pad = F, size = .5) +
  labs(x = expression(Delta~AIC), y = "CDF") +
  theme_minimal()+
  scale_x_continuous(
    limits = c(-4, 14),
    #breaks = seq(-4, 12, by = 1)
    breaks = seq(-4, 14, by = 2)
  ) +
  scale_linetype_manual(values = linetypes) + # Apply manual linetype scale
  scale_color_manual(values = colortypes) +
  theme(
    panel.grid.minor.x = element_blank(),  # Remove minor grid lines on the x-axis
    panel.grid.minor.y = element_blank()
  ) +
  #guides(color = guide_legend(title = NULL)) +
  #guides(color = 'none') +
  geom_vline(xintercept = 2, linetype = "dashed", color = "black", size = 0.5) +
  theme(
    #plot.background = element_rect(fill = "white"),
    #panel.background = element_rect(fill = "white"),
    legend.box.background = element_rect(fill = "white"),
    legend.text = element_text(size = 6),
    legend.key.size = unit(1.1, "lines"),
    legend.title = element_text(size = 0),
    legend.position = c(0.15, 0.70)
)

cdf_plot

# Save the plot with a specific width in centimeters
#ggsave("Figs/cdf-sem_AIC.png", plot = cdf_plot , width = 8.7, height = 5, units = "cm")

```


### Plot AIC and BIC results

```{r}

# read the results
cid.bic <- tibble(
  t100=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_bic(., 4, 100),
  t500=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_bic(., 4, 500),
  t1000=get_item(readRDS(file='R/data/out/lik-CID4-phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_bic(., 4, 1000)
)

sem.bic <- tibble(
  t100=get_item(readRDS( file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-100tips-100tr.RDS'), 'lnLik') %>% get_bic(., 2, 100),
  t500=get_item(readRDS( file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-500tips-100tr.RDS'), 'lnLik') %>% get_bic(., 2, 500),
  t1000=get_item(readRDS(file='R/data/out/SEM/best_Ln-SEM8_phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_bic(., 2, 1000)
)


```



Get dBIC

```{r}

# get max delta aic
del.bic <- cid.bic-sem.bic
apply(del.bic, 2, max)
apply(del.bic, 2, min)

# Calculate ECDF for each column
ecdf_del_bic <- del.bic %>%
  summarise(
    t100_ecdf = ecdf(t100),
    t500_ecdf = ecdf(t500),
    t1000_ecdf = ecdf(t1000)
  )


# Create a data frame for plotting
ecdf_data_bic <- data.frame(
  Value = c(del.bic$t100, del.bic$t500, del.bic$t1000),
  Group = rep(c("100 tips", "500 tips", "1000 tips"), each = nrow(del.bic))
)


```

ECDF: Proportion of trials with dBIC > 2.

```{r}

1-ecdf_del_bic$t100_ecdf(2)
1-ecdf_del_bic$t500_ecdf(2)
1-ecdf_del_bic$t1000_ecdf(2)

```


Combine AIC and BIC

```{r}
# Bind the two dataframes by rows
ecdf_data_combined <- rbind(mutate(ecdf_data, IC = 'AIC'), mutate(ecdf_data_bic, IC = 'BIC'))

#ecdf_data_combined %>% filter(is.finite(Value))
```



#### Plot

```{r}


# Define linetypes for each group
linetypes <- c("100 tips" = "dotted", "500 tips" = "dashed", "1000 tips" = "solid")
colortypes <- c("100 tips" = "#619CFF", "500 tips" = "#00BA38", "1000 tips" = "#F8766D")
#linetypes <- c("100 tips" = "dotted", "500 tips" = "dashed")

# Plot the ECDFs for each group with varying linetypes
cdf_plot <- ggplot() +
  stat_ecdf(data = ecdf_data_combined, aes(x = Value, linetype = Group, color=IC), geom = "step", pad = F, size = .5) +
  labs(x = expression(Delta*AIC~"&"~Delta*BIC), y = "CDF") +
  theme_minimal()+
  scale_x_continuous(
    limits = c(-4, 26),
    breaks = c(c(-4, 0, 2),seq(4, 26, by = 4))
    #breaks = seq(-4, 26, by = 4)
  ) +
  scale_linetype_manual(values = linetypes, name = NULL) + # Apply manual linetype scale
  scale_color_discrete(guide = "none", name = NULL) +
  #scale_color_manual(values = colortypes) +
  theme(
    panel.grid.minor.x = element_blank(),  # Remove minor grid lines on the x-axis
    panel.grid.minor.y = element_blank()
  ) +
  geom_vline(xintercept = 2, linetype = "dashed", color = "black", size = 0.5) +
  theme(
    legend.box.background = element_rect(fill = "white"),
    legend.text = element_text(size = 6),
    legend.key.size = unit(1.1, "lines"),
    #legend.position = c(0.15, 0.70),
    legend.position = c(0.85, 0.35),
    legend.margin = margin(0, 0, 0, 0),    
    legend.box.margin = margin(-5, 4, 0, 2) 
    #legend.spacing = unit(0.01, "cm") 
)

cdf_plot

# Save the plot with a specific width in centimeters
#ggsave("Figs/cdf.png", plot = cdf_plot , width = 8.7, height = 5, units = "cm")

```


### Unique Q matrices

A high number of unique Q matrices indicates that, by chance, a different model is almost always selected as the best.

```{r}
# 100 tips
Q_100 <- readRDS( file='R/data/out/SEM/best_Qs-SEM8_phy_CID4-100tips-100tr.RDS')
N_unique_Qs(Q_100)
# [1] 92

# 500 tips
Q_500 <- readRDS( file='R/data/out/SEM/best_Qs-SEM8_phy_CID4-500tips-100tr.RDS')
N_unique_Qs(Q_500)
# [1] 86

# 1000 tips
Q_1000 <- readRDS( file='R/data/out/SEM/best_Qs-SEM8_phy_CID4-1000tips-100tr.RDS')
N_unique_Qs(Q_1000)
# [1] 68
```


Here, we calculate proportions of unique Qs in models with dAIC >2 (SEM8 is preffered)

```{r}
#del.aic$t100, del.aic$t500, del.aic$t1000

# 100 tips
Q_100 <- readRDS( file='R/data/out/SEM/best_Qs-SEM8_phy_CID4-100tips-100tr.RDS')
index=which(del.aic$t100 >2 )
Q_selected=Q_100[index]
Q_unique=N_unique_Qs(Q_selected)
length(Q_selected)/length(Q_unique)

# 500 tips
Q_500 <- readRDS( file='R/data/out/SEM/best_Qs-SEM8_phy_CID4-500tips-100tr.RDS')
index=which(del.aic$t500 >2 )
Q_selected=Q_500[index]
Q_unique=N_unique_Qs(Q_selected)
length(Q_selected)/length(Q_unique)

# 1000 tips
Q_1000 <- readRDS( file='R/data/out/SEM/best_Qs-SEM8_phy_CID4-1000tips-100tr.RDS')
index=which(del.aic$t1000 >2 )
Q_selected=Q_1000[index]
Q_unique=N_unique_Qs(Q_selected)
length(Q_selected)/length(Q_unique)


```



### Checking if the selected models are irreducible and trait-dependent

Checking if Qs are lumpable, if they are not lumpable,  it means they are irreducible models. In our analyses, all selected best models are irreducible.

```{r}

# all Qs are not lumpable: is_slumpable() -> False

part_scheme=list(c(1, 2), c(3,4), c(5,6), c(7,8))
lapply(Q_100, function(x) is_slumpable(x, part_scheme)) %>% unlist %>% any()
lapply(Q_500, function(x) is_slumpable(x, part_scheme)) %>% unlist %>% any()
lapply(Q_1000, function(x) is_slumpable(x, part_scheme)) %>% unlist %>% any()
```

All SEM models tested are trait-dependent due to their lambda tensor. However, we can also examine how many of the selected models have a Q matrix that implies trait-dependence as well. Note a small proportion of Qs in each analysis are trait-independent. However, typically these Q's imply that trait evolution is diversity-dependent though.

```{r}

# the function is_trait_INdependent() checks if Q is trait-INdependent
lapply(Q_100, function(x) is_trait_INdependent(x)) %>% unlist %>% sum()
lapply(Q_500, function(x) is_trait_INdependent(x)) %>% unlist %>% sum()
lapply(Q_1000, function(x) is_trait_INdependent(x)) %>% unlist %>% sum()

```

### Plot Q's

```{r}
Qfocal <- Q_100[[1]]
Qfocal
# Plot the graph
g <- create_graph_from_matrix(Qfocal)
unidirectional_edges <- is_unidirectional(g)
plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))
```
