---
#title: "Congruence under varying Sampling Fraction"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
params:
  output_dir: "/output"
---

# Congruence under varying sampling fraction

Hrein, we show that congruence holds under varying sampling fraction using the previous. 

```{r,  include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here(), cache = TRUE)
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
source('R/utils/dependencies.R')
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```

## Read in data

```{r}
NSIM = 10
phy <- readRDS(file='R/data/sim/phy_CID4-10tr.RDS')
```


## Setup Sampling Fraction

```{r}
s1 <- 0.6
s2 <- 0.3

sam.fr4 = list(
  c(1-s1, 1-s2, 1-s1, 1-s2,   s1, 0, s1, 0),
  c(1-s1, 1-s2, 1-s1, 1-s2,   0, s2, 0, s2)
)

sam.fr8 = list(
  c(1-s1, 1-s2, 1-s1, 1-s2, 1-s1, 1-s2, 1-s1, 1-s2,   s1,0,s1,0,s1,0,s1,0),
  c(1-s1, 1-s2, 1-s1, 1-s2, 1-s1, 1-s2, 1-s1, 1-s2,   0,s2,0,s2,0,s2,0,s2)
)
```



## Setting-up Q matrices for inference

### CID4

```{r}
# trait
Q_t <- initQ(c(0,1), c(.1, .1))

# diversification regime
Q_r2 <- initQ(c('A','B'), c(.1,.1))


# speciation rates for tracking the order
La4 <- diag(c(-0.1,-0.05),2)
La4 = La4 %x% diag(1,2)
# order according to regimes
Q_cid4.r <- amaSMM(Q_r2, Q_t)

# order according to trait
v=c(1,3, 2,4)
Q_cid4.t <- Q_cid4.r[v,v]

print(Q_cid4.r)
print(La4)
print(Q_cid4.t)
print(La4[v,v])

```


### CID8 (congruent to CID4)

```{r}
# diversification regime
#Q_r4 <- initQ(c('a','b', 'c', 'd'), c(.01,.01))

Q_r4 <-amaSMM(Q_t, Q_t) 
colnames(Q_r4) <- rownames(Q_r4) <- c('a','b', 'c', 'd')

# speciation rates for tracking the order
La8 <- diag(c(-0.1,-0.05),2)
La8 = La8 %x% diag(1,4)

# order according to regimes
Q_cid8.r <- amaSMM(Q_r4, Q_t) 

# order according to trait
v=c(1,3,5,7, 2,4,6,8)
Q_cid8.t <- Q_cid8.r[v,v]

print(Q_cid8.r)
print(La8)
print(Q_cid8.t)
print(La8[v,v])

```

### COR8-C (correlated evolution, congruent to CID4)

```{r}
# like CID8 but with correlation
Q_cor8_C.t <- Q_cid8.t

bl <- matrix(c(0,.1,.1,0),2,2)
Q_cor8_C.t[1:2,5:6] <- bl
Q_cor8_C.t[3:4,7:8] <- bl
Q_cor8_C.t[5:6,1:2] <- bl
Q_cor8_C.t[7:8,3:4] <- bl

v=c(1,5, 2,6, 3,7, 4,8)
Q_cor8_C.r <- Q_cor8_C.t[v,v]

print(Q_cor8_C.r)
#print(La8)
print(Q_cor8_C.t)
#print(La8[v,v])

```

### COR8-NC (Non-correlated evolution, congruent to CID4)

```{r}
# like CID8 but not congruent
Q_cor8_NC.t <- Q_cid8.t

bl <- matrix(c(0,0,.1,0),2,2)
Q_cor8_NC.t[1:2,5:6] <- bl
Q_cor8_NC.t[3:4,7:8] <- bl
Q_cor8_NC.t[5:6,1:2] <- bl
Q_cor8_NC.t[7:8,3:4] <- bl

v=c(1,5, 2,6, 3,7, 4,8)
Q_cor8_NC.r <- Q_cor8_NC.t[v,v]

print(Q_cor8_NC.r)
#print(La8)
print(Q_cor8_NC.t)
#print(La8[v,v])

```


## Irreducible model

### CID4 (4 pars)

```{r}


Args <- list(
  Nstates = 4L,
  y = sam.fr4)

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(4)
#args$arrays
#args$pars
#length(args$pars)
reoder_lambdas(args$arrays, c(1,3, 2,4))

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.05
pars.hc['lam333'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <-0.1
qs <- extract_off_diagonal(Q_cid4.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs



zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid4.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam333 ~ lam222)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r message=FALSE}

CID4 <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- c(1/4, 1/4, 1/4, 1/4)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,   strict=TRUE, control=list(backend = "gslode"), newArgs)
  lik.const <- constrain(lik.c, formulae = f.list)
  arg.const <- argnames(lik.const)
  starting.point <- pars.hc[arg.const]
  CID4[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
}
```

Some helper functions to assess the inference parameters.

```{r}
get_item(CID4, 'lnLik')

```

## Inferences using parameters from CID4

### CID8 (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = sam.fr8)

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#reoder_lambdas(args$arrays, c(1,3,5,7, 2,4,6,8))
```

Run inference

```{r message=FALSE}


CID8.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- est['lam000']
  pars.hc['lam111'] <- est['lam000']
  pars.hc['lam222'] <- est['lam000']
  pars.hc['lam333'] <- est['lam000']
  pars.hc['lam444'] <- est['lam222']
  pars.hc['lam555'] <- est['lam222']
  pars.hc['lam666'] <- est['lam222']
  pars.hc['lam777'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cid8.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  #pars.hc
  #pars_to_arrays(pars.hc, 8)
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  CID8.pars[[i]] <- lik.c(pars.hc, intermediates=T, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Compare Ln

```{r}
get_item(CID4, 'lnLik')
CID8.pars %>% unlist

```

### COR8-C (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = sam.fr8)

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args

```

Run inference

```{r message=FALSE}


COR8_C.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- est['lam000']
  pars.hc['lam111'] <- est['lam000']
  pars.hc['lam222'] <- est['lam000']
  pars.hc['lam333'] <- est['lam000']
  pars.hc['lam444'] <- est['lam222']
  pars.hc['lam555'] <- est['lam222']
  pars.hc['lam666'] <- est['lam222']
  pars.hc['lam777'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_C.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  #v=c(1,3,5,7, 2,4,6,8)
  #pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  COR8_C.pars[[i]] <- lik.c(pars.hc, root=ROOT.GIVEN, root.p=root, condition.surv=T, intermediates=F)
  
}

```

Compare Ln

```{r}
get_item(CID4, 'lnLik')
COR8_C.pars %>% unlist

```

### COR8-NC (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = sam.fr8)

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args

```

Run inference

```{r message=FALSE}


COR8_NC.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  pars.hc['lam000'] <- est['lam000']
  pars.hc['lam111'] <- est['lam000']
  pars.hc['lam222'] <- est['lam000']
  pars.hc['lam333'] <- est['lam000']
  pars.hc['lam444'] <- est['lam222']
  pars.hc['lam555'] <- est['lam222']
  pars.hc['lam666'] <- est['lam222']
  pars.hc['lam777'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_NC.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  # v=c(1,3,5,7, 2,4,6,8)
  # pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  COR8_NC.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Some helper function to assess the inference parameters

```{r}
get_item(CID4, 'lnLik')
COR8_NC.pars %>% unlist

```

### CLA8-C (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = sam.fr8)

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
# reoder_lambdas(args$arrays, c(1,3,5,7, 2,4,6,8))
```

Run inference

```{r message=FALSE}


CLA8_C.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  # pars.hc['lam000'] <- est['lam000']
  # pars.hc['lam111'] <- est['lam000']
  # pars.hc['lam222'] <- est['lam000']
  # pars.hc['lam333'] <- est['lam000']
  # pars.hc['lam444'] <- est['lam222']
  # pars.hc['lam555'] <- est['lam222']
  # pars.hc['lam666'] <- est['lam222']
  # pars.hc['lam777'] <- est['lam222']
  pars.hc['lam022'] <- est['lam000']
  pars.hc['lam133'] <- est['lam000']
  pars.hc['lam202'] <- est['lam000']
  pars.hc['lam311'] <- est['lam000']
  pars.hc['lam466'] <- est['lam222']
  pars.hc['lam577'] <- est['lam222']
  pars.hc['lam646'] <- est['lam222']
  pars.hc['lam757'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_C.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  # v=c(1,3,5,7, 2,4,6,8)
  # pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  CLA8_C.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Some helper function to assess the inference parameters

```{r}
get_item(CID4, 'lnLik')
CLA8_C.pars %>% unlist

```

### CLA8-NC (4 pars)

```{r}


Args <- list(
  Nstates = 8L,
  y = sam.fr8)

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#v=c(1,3,5,7, 2,4,6,8)
#reoder_lambdas(args$arrays, c(1,3,5,7, 2,4,6,8))
```

Run inference

```{r message=FALSE}


CLA8_NC.pars <- list()
i=1
for (i in 1:NSIM){
  print(paste0('Working on: ', i))
  #-----------
  est <- get_item(CID4[i], 'par')
  pars.hc <- rep(0, length(args$pars))
  names(pars.hc) <- args$pars
  # lam004, lam116, lam422 violate lumpability
  pars.hc['lam004'] <- est['lam000']
  pars.hc['lam116'] <- est['lam000']
  pars.hc['lam202'] <- est['lam000']
  pars.hc['lam311'] <- est['lam000']
  pars.hc['lam422'] <- est['lam222']
  pars.hc['lam577'] <- est['lam222']
  pars.hc['lam646'] <- est['lam222']
  pars.hc['lam757'] <- est['lam222']
  pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <- est['mu0']
  qs <- extract_off_diagonal(Q_cor8_C.r)
  qsl=length(qs)
  qs[qs==0.1] <- est['q01']
  qs[qs==0.01] <- est['q02']/2
  pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs
  
  # pars.hc
  # pars_to_arrays(pars.hc, 8)
  # v=c(1,3,5,7, 2,4,6,8)
  # pars_to_arrays(pars.hc, 8)$Q[v,v]
  
  #------
  tree <- phy[[i]]
  states<- tree$tip.state
  states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
  root <- rep(1/8, 8)
  
  lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
  CLA8_NC.pars[[i]] <- lik.c(pars.hc, intermediates=F, root=ROOT.GIVEN, root.p=root, condition.surv=T)
  
}

```

Some helper function to assess the inference parameters

```{r}
get_item(CID4, 'lnLik')
CLA8_NC.pars %>% unlist

```


## Results 

### Inferences using parameters from CID4

#### Likelihood table

```{r}


cid4 = get_item(CID4, 'lnLik')

cor.pars <- tibble(
"N" = c(1:NSIM),
"cid4" = cid4,
"CID8.pars"= CID8.pars %>% unlist,
"COR8-C.pars"= COR8_C.pars %>% unlist,
"CLA8-C.pars" = CLA8_C.pars %>% unlist,
"COR8-NC.pars" = COR8_NC.pars %>% unlist,
"CLA8-NC.pars" = CLA8_NC.pars %>% unlist,
)

kable(cor.pars, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```


#### Mean absolute error

```{r}

dcor.pars <- tibble(
"cid4" = cid4,
"CID8.pars"= CID8.pars %>% unlist,
"COR8-C.pars"= COR8_C.pars %>% unlist,
"CLA8-C.pars" = CLA8_C.pars %>% unlist,
"COR8-NC.pars" = COR8_NC.pars %>% unlist,
"CLA8-NC.pars" = CLA8_NC.pars %>% unlist,
)
cid4t <- cid4
dcor.pars <- dcor.pars %>%
  mutate_all(~ abs(cid4t - .) )

column_means <- dcor.pars %>%
  summarise_all(~ mean(.))

# Data
delta_values <- tibble(
  Delta = names(column_means),
  Value = column_means %>% slice(1) %>% unlist()
)

kable(column_means, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Create the plot
ggplot(delta_values, aes(x = Delta, y = Value, label = sprintf("%.2e", Value))) +
  geom_col(fill = c("blue", "blue", "blue", "blue", "red",  "red")) +
  geom_text(vjust = -0.5, angle = 45) +
  labs(title = "Comparison of Ln Differences", x = "Deltas", y = "Magnitude") +
  theme_minimal()


```


