---
#title: " Congruence between independent and dependent SSEs"
author: "Sergei Tarasov"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
params:
  output_dir: "/output"
---

# Topological Bridges

Here, we show that CID4 may be probabilistically closer to irreducible trait-dependent models than to other irreducible CID models.


```{r,  include=FALSE}
library(DT)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
knitr::opts_knit$set(root.dir = here::here(), cache = TRUE)
```

```{r message=FALSE}

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##  ~ installations and dependencies  ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
source('R/utils/dependencies.R')
source('R/hiclasse/HiClaSSE-R.R') # pure R implementation of HiCLaSSE
source('R/hiclasse/HiClaSSE_cpp.R') # fast implementation


```

## Read in data

```{r}
NSIM = 100
#phy <- readRDS(file='R/data/sim/phy_CID4-500tips-100tr.RDS')
files_base <- c('phy_CID4-100tips-100tr', 'phy_CID4-500tips-100tr', 'phy_CID4-1000tips-100tr')

```



## Setting-up Q matrices for inference

### CID4

```{r}
# trait
Q_t <- initQ(c(0,1), c(.1, .1))

# diversification regime
Q_r2 <- initQ(c('A','B'), c(.1,.1))


# speciation rates for tracking the order
La4 <- diag(c(-0.1,-0.05),2)
La4 = La4 %x% diag(1,2)
# order according to regimes
Q_cid4.r <- amaSMM(Q_r2, Q_t)

# order according to trait
v=c(1,3, 2,4)
Q_cid4.t <- Q_cid4.r[v,v]

print(Q_cid4.r)
print(La4)
print(Q_cid4.t)
print(La4[v,v])

```



### CID8 (congruent to CID4)

```{r}
# diversification regime
#Q_r4 <- initQ(c('a','b', 'c', 'd'), c(.01,.01))

Q_r4 <-amaSMM(Q_t, Q_t) 
colnames(Q_r4) <- rownames(Q_r4) <- c('a','b', 'c', 'd')

# speciation rates for tracking the order
La8 <- diag(c(-0.1,-0.05),2)
La8 = La8 %x% diag(1,4)

# order according to regimes
Q_cid8.r <- amaSMM(Q_r4, Q_t) 

# order according to trait
v=c(1,3,5,7, 2,4,6,8)
Q_cid8.t <- Q_cid8.r[v,v]

print(Q_cid8.r)
print(La8)
print(Q_cid8.t)
print(La8[v,v])

```
### CID8-NC

```{r}
# diversification regime
Q_r4 <- initQ(c('a','b', 'c', 'd'), c(.1,.1))

Q_r4[1,3] <- 0
Q_r4[1,4] <- 0
Q_r4[2,3] <- 0

print(Q_r4)

#Q_r4 <-amaSMM(Q_t, Q_t) 
#colnames(Q_r4) <- rownames(Q_r4) <- c('a','b', 'c', 'd')


# order according to regimes
Q_cid8_NC.r <- amaSMM(Q_r4, Q_t) 

# order according to trait
v=c(1,3,5,7, 2,4,6,8)
Q_cid8_NC.t <- Q_cid8_NC.r[v,v]

print(Q_cid8_NC.r)
print(Q_cid8_NC.t)

```

```{r, eval=FALSE}

# Plot the graph
g <- create_graph_from_matrix(Q_cid8_NC.t)
unidirectional_edges <- is_unidirectional(g)
plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))

```


### COR8-C (correlated evolution, congruent to CID4)

```{r, eval=FALSE}
# like CID8 but with correlation
Q_cor8_C.t <- Q_cid8.t


Q_cor8_C.t['b0', 'd0'] <- 0
Q_cor8_C.t['b0', 'c0'] <- 0.1

v=c(1,5, 2,6, 3,7, 4,8)
Q_cor8_C.r <- Q_cor8_C.t[v,v]

print(Q_cor8_C.r*10)
#print(La8)
print(Q_cor8_C.t*10)
#print(La8[v,v])

# Plot the graph
g <- create_graph_from_matrix(Q_cor8_C.t)
unidirectional_edges <- is_unidirectional(g)
plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))

```

### COR8-NC (Correlated evolution, noncongruent to CID4)

```{r}
# not congruent
Q_cor8_NC.t <- Q_cid8.t

Q_cor8_NC.t['b0', 'd0'] <- 0
Q_cor8_NC.t['b0', 'c0'] <- 0

v=c(1,5, 2,6, 3,7, 4,8)
Q_cor8_NC.r <- Q_cor8_NC.t[v,v]

print(Q_cor8_NC.r*10)
#print(La8)
print(Q_cor8_NC.t*10)
#print(La8[v,v])

# Plot the graph
g <- create_graph_from_matrix(Q_cor8_NC.t)
unidirectional_edges <- is_unidirectional(g)
#plot(g, vertex.label = V(g)$name, vertex.size = 30, edge.color = ifelse(unidirectional_edges, "red", "black"))

```


## Maximum Likelihood. 

### CID4: true model

```{r, eval=FALSE}


Args <- list(
  Nstates = 4L,
  y = list(
    c(0,0,0,0, 1,0,1,0),
    c(0,0,0,0, 0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(4)
#args$arrays
#args$pars
#length(args$pars)
reoder_lambdas(args$arrays, c(1,3, 2,4))

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.05
pars.hc['lam333'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <-0.1
qs <- extract_off_diagonal(Q_cid4.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,4)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid4.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam333 ~ lam222)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r message=FALSE, eval=FALSE}

#base="phy_CID4-100tips-100tr"
for (base in files_base[1:3]){
  file=file.path("R/data/sim", paste0(base, '.RDS'))
  print(paste0('Reading file: ',file))
  phy <- readRDS(file=file)

  CID4 <- list()
  #i=1
  for (i in 1:NSIM){
    print(paste0('Working on: ', i))
    tree <- phy[[i]]
    states<- tree$tip.state
    states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
    root <- c(1/4, 1/4, 1/4, 1/4)
    
    lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
    lik.const <- constrain(lik.c, formulae = f.list)
    arg.const <- argnames(lik.const)
    starting.point <- pars.hc[arg.const]
    starting.point['lam000'] <- 0.3
    starting.point['lam222'] <- 0.1
    starting.point['mu0'] <- 0.01
    starting.point['q01'] <- 0.2
    CID4[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
  }
  
  file_out=paste0('R/data/out/topo-bridges/CID4-', base, '.RDS')
  saveRDS(CID4, file= file_out)

}

```



### CID8-NC 

```{r, eval=FALSE}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.1
pars.hc['lam333'] <- 0.1
pars.hc['lam444'] <- 0.05
pars.hc['lam555'] <- 0.05
pars.hc['lam666'] <- 0.05
pars.hc['lam777'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cid8_NC.r)
qsl=length(qs)
#unique(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid8_NC.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam222 ~ lam000, lam333 ~ lam000,   
            lam555 ~ lam444, lam666 ~ lam444, lam777 ~ lam444)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```



Run inference

```{r, eval=FALSE}
for (base in files_base[1:3]){
  file=file.path("R/data/sim", paste0(base, '.RDS'))
  print(paste0('Reading file: ',file))
  phy <- readRDS(file=file)
  
  CID8_NC <- list()
  i=1
  for (i in 1:NSIM){
    print(paste0('Working on: ', i))
    tree <- phy[[i]]
    states<- tree$tip.state
    states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
    root <- rep(1/8, 8)
    
    lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
    lik.const <- constrain(lik.c, formulae = f.list)
    arg.const <- argnames(lik.const)
    starting.point <- pars.hc[arg.const]
    #starting.point <- pars.hc[arg.const]
    starting.point['lam000'] <- 0.3
    starting.point['lam444'] <- 0.1
    starting.point['mu0'] <- 0.01
    starting.point['q01'] <- 0.1
    
    CID8_NC[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
  }
  
  file_out=paste0('R/data/out/topo-bridges/CID8_NC-', base, '.RDS')
  saveRDS(CID8_NC, file= file_out)

}
```


### CID8a-NC

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args$arrays
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.1
pars.hc['lam333'] <- 0.1

pars.hc['lam204'] <- 0.1
pars.hc['lam315'] <- 0.1

pars.hc['lam444'] <- 0.05
pars.hc['lam555'] <- 0.05
pars.hc['lam666'] <- 0.05
pars.hc['lam777'] <- 0.05

# pars.hc['lam602'] <- 0.05
# pars.hc['lam713'] <- 0.05

pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cid8_NC.r)
qsl=length(qs)
#unique(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cid8_NC.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam222 ~ lam000, lam333 ~ lam000, 
            lam204 ~ lam000, lam315 ~ lam000,
            #lam602 ~ lam444, lam713 ~ lam444,
            lam555 ~ lam444, lam666 ~ lam444, lam777 ~ lam444)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Let's check the model rates
```{r, eval=FALSE}
# a0 a1 b0 b1 c0 c1 d0 d1
ar=names2array(args$arrays, c('a0', 'a1', 'b0', 'b1', 'c0', 'c1', 'd0', 'd1'))
ar.re=reoder_lambdas(ar, c(1,3,5,7, 2,4,6,8))
#ar.re$lam.tensor


# rates that will be used to setup model
element <- c( 
             'lam204',
             'lam315',
             'lam602',
             'lam713'
             )
find_elements(ar.re$lam.tensor, element)
```

Run inference

```{r, eval=FALSE}
for (base in files_base[1:3]){
  file=file.path("R/data/sim", paste0(base, '.RDS'))
  print(paste0('Reading file: ',file))
  phy <- readRDS(file=file)

  CID8a_NC <- list()
  #i=1
  for (i in 1:NSIM){
    print(paste0('Working on: ', i))
    tree <- phy[[i]]
    states<- tree$tip.state
    states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
    root <- rep(1/8, 8)
    
    lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
    lik.const <- constrain(lik.c, formulae = f.list)
    arg.const <- argnames(lik.const)
    starting.point <- pars.hc[arg.const]
    #starting.point <- pars.hc[arg.const]
    starting.point['lam000'] <- 0.3
    starting.point['lam444'] <- 0.1
    starting.point['mu0'] <- 0.01
    starting.point['q01'] <- 0.1
    
    CID8a_NC[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
  }
  
  file_out=paste0('R/data/out/topo-bridges/CID8a_NC-', base, '.RDS')
  saveRDS(CID8a_NC, file= file_out)

}
```



### COR8-C

```{r, eval=FALSE}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.1
pars.hc['lam333'] <- 0.1
pars.hc['lam444'] <- 0.05
pars.hc['lam555'] <- 0.05
pars.hc['lam666'] <- 0.05
pars.hc['lam777'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cor8_C.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cor8_C.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam222 ~ lam000, lam333 ~ lam000,   
            lam555 ~ lam444, lam666 ~ lam444, lam777 ~ lam444)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```



Run inference

```{r, eval=FALSE}
for (base in files_base[1:3]){
  file=file.path("R/data/sim", paste0(base, '.RDS'))
  print(paste0('Reading file: ',file))
  phy <- readRDS(file=file)
  
  COR8_C <- list()
  i=1
  for (i in 1:NSIM){
    print(paste0('Working on: ', i))
    tree <- phy[[i]]
    states<- tree$tip.state
    states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
    root <- rep(1/8, 8)
    
    lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
    lik.const <- constrain(lik.c, formulae = f.list)
    arg.const <- argnames(lik.const)
    starting.point <- pars.hc[arg.const]
    starting.point['lam000'] <- 0.3
    starting.point['lam444'] <- 0.1
    starting.point['mu0'] <- 0.01
    starting.point['q01'] <- 0.2
    COR8_C[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
  }
  
  file_out=paste0('R/data/out/topo-bridges/COR8_C-', base, '.RDS')
  saveRDS(COR8_C, file= file_out)

}


```



### COR8-NC

```{r}


Args <- list(
  Nstates = 8L,
  y = list(
    c(0,0,0,0,0,0,0,0, 1,0,1,0,1,0,1,0),
    c(0,0,0,0,0,0,0,0, 0,1,0,1,0,1,0,1)
  ))

newArgs <- makeArgs(Args)
#printArgsGlobal()
args <- argnames_HiClaSSE(8)
#args
#args$pars
#length(args$pars)

pars.hc <- rep(0, length(args$pars))
names(pars.hc) <- args$pars
pars.hc['lam000'] <- 0.1
pars.hc['lam111'] <- 0.1
pars.hc['lam222'] <- 0.1
pars.hc['lam333'] <- 0.1
pars.hc['lam444'] <- 0.05
pars.hc['lam555'] <- 0.05
pars.hc['lam666'] <- 0.05
pars.hc['lam777'] <- 0.05
pars.hc['mu0'] <-pars.hc['mu1'] <- pars.hc['mu2'] <- pars.hc['mu3'] <- pars.hc['mu4']<- pars.hc['mu5']<- pars.hc['mu6']<- pars.hc['mu7'] <-0.1
qs <- extract_off_diagonal(Q_cor8_NC.r)
qsl=length(qs)
pars.hc[c(length(pars.hc)-qsl+1):length(pars.hc)] <- qs

#pars.hc
#pars_to_arrays(pars.hc,8)
#args

zero.constr <- formulas_zero_pars(pars.hc)
f.qs <- assign_classes_pairwise(Q_cor8_NC.r, args$arrays$Q)
f.mu <- c(mu1 ~ mu0, mu2 ~ mu0, mu3 ~ mu0, mu4 ~ mu0, mu5 ~ mu0, mu6 ~ mu0, mu7 ~ mu0)
f.lams <- c(lam111 ~ lam000, lam222 ~ lam000, lam333 ~ lam000,   
            lam555 ~ lam444, lam666 ~ lam444, lam777 ~ lam444)
f.list<- c(zero.constr,  f.qs, f.mu, f.lams)

```

Run inference

```{r, eval=FALSE}
for (base in files_base[1:3]){
  file=file.path("R/data/sim", paste0(base, '.RDS'))
  print(paste0('Reading file: ',file))
  phy <- readRDS(file=file)
  
  COR8_NC <- list()
  i=1
  for (i in 1:NSIM){
    print(paste0('Working on: ', i))
    tree <- phy[[i]]
    states<- tree$tip.state
    states<- mapvalues(states, from = c("0", "1", "2", "3"), to=c(0, 1, 0, 1) )
    root <- rep(1/8, 8)
    
    lik.c <- make.HiClasse_cpp(tree, states,  sampling.f=NULL,  strict=TRUE, control=list(backend = "gslode"), newArgs)
    lik.const <- constrain(lik.c, formulae = f.list)
    arg.const <- argnames(lik.const)
    # starting.point <- pars.hc[arg.const]
    starting.point <- get_item(COR8_C[i], 'par')
    COR8_NC[[i]] <- find.mle(lik.const, starting.point, method="subplex", keep.func=F, root=ROOT.GIVEN, root.p=root, condition.surv=TRUE)
  }
  
  file_out=paste0('R/data/out/topo-bridges/COR8_NC-', base, '.RDS')
  saveRDS(COR8_NC, file= file_out)

}
```



## Results

```{r}
`%+%` <- function(a, b) {
  paste0(a, b)
}

'a' %+% 'b'
```


```{r}

#CID8_NC[[1]]$par

get_item(readRDS(file='R/data/out/topo-bridges/COR8_NC-phy_CID4-1000tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean

Ntips=100
t100 <- c(
CID4=get_item(readRDS(file='R/data/out/topo-bridges/CID4-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
COR8_C=get_item(readRDS(file='R/data/out/topo-bridges/COR8_C-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
COR8_NC=get_item(readRDS(file='R/data/out/topo-bridges/COR8_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
CID8_NC=get_item(readRDS(file='R/data/out/topo-bridges/CID8_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
CID8a_NC=get_item(readRDS(file='R/data/out/topo-bridges/CID8a_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean
)

Ntips=500
t500 <- c(
CID4=get_item(readRDS(file='R/data/out/topo-bridges/CID4-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
COR8_C=get_item(readRDS(file='R/data/out/topo-bridges/COR8_C-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
COR8_NC=get_item(readRDS(file='R/data/out/topo-bridges/COR8_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
CID8_NC=get_item(readRDS(file='R/data/out/topo-bridges/CID8_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
CID8a_NC=get_item(readRDS(file='R/data/out/topo-bridges/CID8a_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean
)

Ntips=1000
t1000 <- c(
CID4=get_item(readRDS(file='R/data/out/topo-bridges/CID4-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
COR8_C=get_item(readRDS(file='R/data/out/topo-bridges/COR8_C-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
COR8_NC=get_item(readRDS(file='R/data/out/topo-bridges/COR8_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
CID8_NC=get_item(readRDS(file='R/data/out/topo-bridges/CID8_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean,
CID8a_NC=get_item(readRDS(file='R/data/out/topo-bridges/CID8a_NC-phy_CID4-' %+% Ntips %+% 'tips-100tr.RDS'), 'lnLik') %>% get_aic(., 4) %>% mean
)

aic=rbind(t100=t100, t500=t500, t1000=t1000)
daic=apply(aic, 2, function(x) x-aic[,1]) %>% round(., 3)

```

### dAIC is a proxy to relative KL divergence
```{r}

print(daic)
```




